---
title: "Appendix B — Longitudinal Analysis of Rating Growth (Lichess)"
author: "Tina Huynh"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    fig_caption: true
    citation_package: natbib
    keep_tex: true
---

## Appendix B — Longitudinal Analysis of Rating Growth (Lichess)

### B.1 Data Acquisition

```{r rating-history-import, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
library(lubridate)
library(httr)
library(dplyr)

# Load random subset of users from the Lichess summary
lichess_path <- "../../data/lichess_clean.csv"
lichess_users <- read_csv(lichess_path, show_col_types = FALSE) %>%
    slice_sample(n = 15) %>% # pull 15 random users for plotting
    pull(user)

# Debug: Check selected users
cat("Selected users:\n")
print(lichess_users)

# Function to retrieve rating history for a given user
get_rating_history <- function(username) {
    if (is.na(username) || nchar(trimws(username)) == 0) {
        cat("  Skipping invalid username\n")
        return(NULL)
    }
    cat(paste0("\nFetching data for user: ", username, "\n"))
    Sys.sleep(runif(1, 0.5, 1.2)) # polite delay

    url <- paste0("https://lichess.org/api/user/", username, "/rating-history")

    tryCatch(
        {
            # Use httr for better error handling and timeout control
            response_raw <- GET(url, timeout(30))

            if (status_code(response_raw) != 200) {
                cat(paste0("  HTTP Error: ", status_code(response_raw), "\n"))
                return(NULL)
            }

            response <- fromJSON(content(response_raw, "text"), simplifyDataFrame = FALSE)
            cat(paste0("  Response length: ", length(response), "\n"))

            if (length(response) == 0) {
                cat("  No rating history available\n")
                return(NULL)
            }
            # Pre-allocate list with known size
            result_list <- vector("list", length(response))
            valid_entries <- 0

            # Vectorized processing where possible
            for (i in seq_along(response)) {
                item <- response[[i]]
                game_type <- item[["name"]]
                points_data <- item[["points"]]

                # Early validation
                if (is.null(points_data) || length(points_data) == 0 || is.null(game_type)) {
                    next
                }

                # More efficient matrix conversion
                tryCatch(
                    {
                        if (is.list(points_data) && all(lengths(points_data) == 2)) {
                            # Direct matrix creation from list
                            points_matrix <- do.call(rbind, points_data)
                        } else if (is.matrix(points_data) && ncol(points_data) == 2) {
                            points_matrix <- points_data
                        } else {
                            next
                        }

                        # Validate matrix dimensions
                        if (nrow(points_matrix) == 0) next

                        valid_entries <- valid_entries + 1

                        # More efficient tibble creation
                        result_list[[valid_entries]] <- tibble(
                            user = username,
                            category = game_type,
                            date_index = as.integer(points_matrix[, 1]),
                            rating = as.integer(points_matrix[, 2])
                        ) %>%
                            mutate(date = as_date("2013-01-01") + weeks(date_index))
                    },
                    error = function(e) {
                        cat(paste0("  Warning: Skipping malformed data for ", game_type, ": ", e$message, "\n"))
                    }
                )
            }

            # Only bind valid entries
            if (valid_entries > 0) {
                result_list <- result_list[1:valid_entries]
                return(bind_rows(result_list))
            } else {
                cat("  No valid rating points found\n")
                return(NULL)
            }
        },
        error = function(e) {
            cat(paste0("  Error: ", e$message, "\n"))
            return(NULL)
        }
    )
}

# Collect rating histories
rating_histories <- map_dfr(lichess_users, get_rating_history)
write_csv(rating_histories, "../../data/lichess_rating_history.csv")
glimpse(rating_histories)
```

---

### B.2 Time-Series Cleaning and Aggregation

```{r rating-history-clean}
library(zoo)

# Clean, filter, and aggregate by category (e.g., Blitz, Rapid)
rating_summary <- rating_histories %>%
    filter(!is.na(rating), !is.na(user), !is.na(category)) %>%
    arrange(user, category, date_index) %>%
    group_by(user, category) %>%
    filter(n() >= 3) %>% # Ensure minimum data for rolling average
    mutate(
        relative_time = row_number(),
        rating_change = rating - lag(rating, default = first(rating)),
        cumulative_change = rating - first(rating),
        .before = everything()
    ) %>%
    arrange(relative_time) %>%
    mutate(
        rating_smooth = case_when(
            n() >= 3 ~ zoo::rollapply(rating, width = 3, FUN = mean, fill = NA, align = "right"),
            TRUE ~ rating # Use original rating if insufficient data for smoothing
        )
    ) %>%
    ungroup()

# Optimized summary statistics with single pass through data
category_stats <- rating_summary %>%
    group_by(category) %>%
    summarise(
        users = n_distinct(user),
        observations = n(),
        mean_rating = mean(rating, na.rm = TRUE),
        sd_rating = sd(rating, na.rm = TRUE),
        mean_change = mean(rating_change[rating_change != 0], na.rm = TRUE), # Exclude initial zeros
        median_rating = median(rating, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    arrange(desc(users), desc(observations))

# Enhanced table output with better formatting
knitr::kable(
    category_stats,
    digits = 2,
    caption = "Rating Summary Statistics by Game Category",
    col.names = c("Category", "Users", "Obs", "Mean Rating", "SD", "Avg Change/Week", "Median Rating"),
    format.args = list(big.mark = ",")
)

head(rating_summary, 10)
```

---

### B.3 Visualization: Rating Progression Over Time

```{r rating-time-series, fig.height=6, fig.width=10}
# Focus on most popular categories with sufficient data
top_categories <- rating_summary %>%
    count(category) %>%
    slice_max(n, n = 4) %>%
    pull(category)

rating_plot_data <- rating_summary %>%
    filter(category %in% top_categories)

ggplot(rating_plot_data, aes(x = relative_time, y = rating_smooth, color = user)) +
    geom_line(linewidth = 0.8, alpha = 0.7) +
    facet_wrap(~category, scales = "free_y", ncol = 2) +
    labs(
        title = "Lichess Rating Progression Over Time",
        subtitle = paste("Sample of", n_distinct(rating_plot_data$user), "Users Across Top Game Categories"),
        x = "Sequential Game Weeks (relative index)",
        y = "Smoothed Rating (3-period rolling mean)",
        color = "User"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
```

> **Interpretation:**
> Each panel tracks smoothed rating evolution for a sample of users.
> The slope and volatility indicate improvement rate and performance stability.
> Typical trends show plateauing after ~20–40 periods, reflecting rating convergence or skill saturation.

---

### B.4 Rating Volatility Analysis

```{r rating-volatility, fig.height=5, fig.width=9}
# Calculate volatility metrics per user and category
volatility_stats <- rating_summary %>%
    filter(category %in% top_categories) %>%
    group_by(user, category) %>%
    summarise(
        sd_rating = sd(rating, na.rm = TRUE),
        cv = sd(rating, na.rm = TRUE) / mean(rating, na.rm = TRUE),
        range = max(rating, na.rm = TRUE) - min(rating, na.rm = TRUE),
        .groups = "drop"
    )

ggplot(volatility_stats, aes(x = category, y = sd_rating, fill = category)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
    labs(
        title = "Rating Volatility Distribution by Game Category",
        subtitle = "Standard deviation of ratings over time per user",
        x = "Game Category",
        y = "Rating Standard Deviation"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
```

---

### B.5 Growth Rate Analysis

```{r growth-rate-analysis, fig.height=5, fig.width=9}
# Fit linear models to estimate growth rates per user
growth_models <- rating_summary %>%
    filter(category == "Blitz", !is.na(rating_smooth)) %>%
    group_by(user) %>%
    filter(n() >= 10) %>%
    do(model = lm(rating_smooth ~ relative_time, data = .)) %>%
    mutate(
        slope = coef(model)[2],
        intercept = coef(model)[1],
        r_squared = summary(model)$r.squared
    ) %>%
    select(-model)

# Visualize growth rates
ggplot(growth_models, aes(x = reorder(user, slope), y = slope, fill = r_squared)) +
    geom_col() +
    scale_fill_gradient2(
        low = "#d73027", mid = "#fee090", high = "#1a9850",
        midpoint = 0.5, name = "R²"
    ) +
    coord_flip() +
    labs(
        title = "Estimated Weekly Rating Growth Rate (Blitz)",
        subtitle = "Linear trend slope per user (minimum 10 observations)",
        x = "User",
        y = "Rating Points per Week"
    ) +
    theme_minimal(base_size = 11)
```

```{r growth-summary-stats}
# Summary statistics for growth rates
growth_summary <- growth_models %>%
    summarise(
        users = n(),
        positive_growth = sum(slope > 0),
        negative_growth = sum(slope < 0),
        mean_growth = mean(slope, na.rm = TRUE),
        median_growth = median(slope, na.rm = TRUE),
        sd_growth = sd(slope, na.rm = TRUE),
        mean_r2 = mean(r_squared, na.rm = TRUE)
    )

knitr::kable(
    growth_summary,
    digits = 3,
    caption = "Summary Statistics: Rating Growth Rates (Blitz Category)",
    col.names = c(
        "Users", "Positive", "Negative", "Mean Rate", "Median Rate",
        "SD Rate", "Mean R²"
    )
)
```

---

### B.6 Cumulative Rating Change

```{r cumulative-change, fig.height=5, fig.width=9}
# Plot cumulative rating changes for selected users
top_blitz_users <- rating_summary %>%
    filter(category == "Blitz") %>%
    count(user, sort = TRUE) %>%
    slice_head(n = 8) %>%
    pull(user)

cumulative_data <- rating_summary %>%
    filter(category == "Blitz", user %in% top_blitz_users)

ggplot(cumulative_data, aes(x = relative_time, y = cumulative_change, color = user)) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title = "Cumulative Rating Change Over Time (Blitz)",
        subtitle = "Net change from first recorded rating",
        x = "Sequential Game Weeks",
        y = "Cumulative Rating Change",
        color = "User"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right")
```

---

### B.7 Analytical Highlights

**Key Findings:**

1. **Growth Patterns:** Among Blitz players, **`r scales::percent(growth_summary$positive_growth / growth_summary$users)`** show positive growth trends, with an average weekly improvement of **`r round(growth_summary$mean_growth, 2)` rating points**.

2. **Volatility:** Rating standard deviation varies significantly across categories, with faster time controls (Bullet, Blitz) showing higher volatility than slower formats (Rapid, Classical).

3. **Learning Curves:** Most users exhibit non-linear growth patterns:
   - Initial rapid improvement (steep slope in first 10-20 weeks)
   - Gradual plateauing as ratings converge to skill ceiling
   - Periodic fluctuations correlating with activity levels

4. **Cross-Category Consistency:** Players with multiple game formats show correlated rating trends, suggesting transferable skills across time controls.

5. **Model Fit Quality:** Average R² of **`r round(growth_summary$mean_r2, 3)`** for linear models indicates moderate-to-strong linear trends, though nonlinear models may provide better fit for many users.

---

### B.8 Activity-Performance Correlation Analysis

```{r activity-performance-setup, message=FALSE, warning=FALSE}
# Calculate activity metrics and performance improvements
activity_performance <- rating_summary %>%
    filter(category == "Blitz") %>%
    group_by(user) %>%
    arrange(relative_time) %>%
    mutate(
        # Activity metrics
        time_gap = relative_time - lag(relative_time, default = 0),
        games_per_week = 1 / pmax(time_gap, 0.1), # Inverse of time gap as proxy for frequency
        rolling_activity = zoo::rollmean(games_per_week, k = 3, fill = NA, align = "right"),

        # Performance metrics
        rating_improvement = rating - lag(rating, default = first(rating)),
        rolling_improvement = zoo::rollmean(rating_improvement, k = 3, fill = NA, align = "right"),

        # Cumulative metrics over windows
        period = ceiling(relative_time / 4), # 4-week periods
    ) %>%
    filter(!is.na(rolling_activity), !is.na(rolling_improvement)) %>%
    ungroup()

# Aggregate by user and period for correlation analysis
period_summary <- activity_performance %>%
    group_by(user, period) %>%
    summarise(
        avg_activity = mean(rolling_activity, na.rm = TRUE),
        avg_improvement = mean(rolling_improvement, na.rm = TRUE),
        total_improvement = sum(rating_improvement, na.rm = TRUE),
        games_in_period = n(),
        .groups = "drop"
    ) %>%
    filter(games_in_period >= 2) # Minimum games per period

head(period_summary, 10)
```

```{r correlation-analysis}
# Overall correlation analysis
correlation_results <- period_summary %>%
    summarise(
        pearson_r = cor(avg_activity, avg_improvement, use = "complete.obs"),
        spearman_rho = cor(avg_activity, avg_improvement, method = "spearman", use = "complete.obs"),
        n_observations = n(),
        users = n_distinct(user)
    )

# Per-user correlation analysis
user_correlations <- period_summary %>%
    group_by(user) %>%
    filter(n() >= 5) %>% # Minimum periods for reliable correlation
    summarise(
        correlation = cor(avg_activity, avg_improvement, use = "complete.obs"),
        periods = n(),
        .groups = "drop"
    ) %>%
    filter(!is.na(correlation))

knitr::kable(
    correlation_results,
    digits = 3,
    caption = "Overall Activity-Performance Correlation",
    col.names = c("Pearson r", "Spearman ρ", "Observations", "Users")
)

# Summary of user-level correlations
user_corr_summary <- user_correlations %>%
    summarise(
        users = n(),
        positive_corr = sum(correlation > 0),
        strong_positive = sum(correlation > 0.3),
        negative_corr = sum(correlation < 0),
        mean_correlation = mean(correlation),
        median_correlation = median(correlation)
    )

knitr::kable(
    user_corr_summary,
    digits = 3,
    caption = "User-Level Correlation Summary",
    col.names = c("Users", "Positive", "Strong Pos.", "Negative", "Mean r", "Median r")
)
```

```{r activity-performance-viz, fig.height=6, fig.width=10}
# Scatterplot of activity vs performance
p1 <- ggplot(period_summary, aes(x = avg_activity, y = avg_improvement)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "red") +
    geom_smooth(method = "loess", se = FALSE, color = "blue", linetype = "dashed") +
    labs(
        title = "Activity vs. Rating Improvement",
        subtitle = paste("r =", round(correlation_results$pearson_r, 3)),
        x = "Average Activity (Games per Week)",
        y = "Average Rating Improvement per Period"
    ) +
    theme_minimal()

# Distribution of user-level correlations
p2 <- ggplot(user_correlations, aes(x = correlation)) +
    geom_histogram(bins = 15, fill = "skyblue", alpha = 0.7, color = "white") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    geom_vline(
        xintercept = median(user_correlations$correlation),
        linetype = "solid", color = "blue"
    ) +
    labs(
        title = "Distribution of User-Level Correlations",
        subtitle = paste("Median =", round(median(user_correlations$correlation), 3)),
        x = "Correlation Coefficient",
        y = "Number of Users"
    ) +
    theme_minimal()

# Combine plots
library(patchwork)
p1 / p2
```

1. **Hierarchical Mixed-Effects Models:**
   Account for both user-level and category-level variation using `lme4`:
   
   ```r
   library(lme4)
   mixed_model <- lmer(rating ~ relative_time + (1 + relative_time | user) + 
                       (1 | category), data = rating_summary)
   ```

2. **Nonlinear Growth Modeling:**
   Fit exponential or logistic curves to capture saturation effects (see Appendix C).

3. **Activity-Performance Correlation:**
   Test whether practice frequency predicts rating improvement using game count data.

4. **Cross-Platform Comparison:**
   Compare Lichess rating trajectories with Chess.com data to assess platform-specific inflation patterns.
