## Appendix D — Interactive Dashboard: Chess Performance Analytics

> **Objective:**  
> Provide a single interactive interface that allows exploration of:
> - Rating distributions (FIDE vs Lichess)  
> - Player win-rate trends  
> - Time-series rating progressions  
> - Growth-model parameters (linear + nonlinear)  

---

### D.1 Setup and Required Packages

```{r dashboard-setup, message=FALSE, warning=FALSE}
# Install once if not already
# install.packages(c("flexdashboard","plotly","DT","broom","here","minpack.lm","lme4","zoo"))

library(flexdashboard)
library(plotly)
library(DT)
library(tidyverse)
library(broom)
library(lubridate)
library(minpack.lm) # for nlsLM
library(lme4)
library(zoo)
```

---

### D.2 Load Data

```{r load-data, message=FALSE}
summary_path <- "../../data/player_summary.csv"
lichess_path <- "../../data/lichess_clean.csv"
rating_hist_path <- "../../data/lichess_rating_history.csv"

player_summary <- read_csv(summary_path, show_col_types = FALSE)
lichess_clean <- read_csv(lichess_path, show_col_types = FALSE)
rating_history <- read_csv(rating_hist_path, show_col_types = FALSE)

# Preview data structures
glimpse(player_summary)
glimpse(lichess_clean)
glimpse(rating_history)
```

---

### D.3 Interactive Rating Distribution Comparison

```{r rating-distribution, fig.height=6, fig.width=9}
# Prepare long format for FIDE vs Lichess comparison
rating_comparison <- player_summary %>%
  select(rating_group, fide_avg_rating_standard, lichess_avg_rating) %>%
  pivot_longer(
    cols = c(fide_avg_rating_standard, lichess_avg_rating),
    names_to = "source",
    values_to = "rating"
  ) %>%
  mutate(source = recode(source,
    fide_avg_rating_standard = "FIDE (Standard)",
    lichess_avg_rating = "Lichess (Online)"
  ))

# Create interactive plotly chart
p1 <- ggplot(rating_comparison, aes(x = rating_group, y = rating, fill = source, 
                                     text = paste0("Rating: ", round(rating, 0), 
                                                  "<br>Group: ", rating_group,
                                                  "<br>Source: ", source))) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("#2E86AB", "#E67E22")) +
  labs(
    title = "Average Player Ratings: FIDE vs Lichess",
    x = "Rating Group",
    y = "Average Rating",
    fill = "Data Source"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p1, tooltip = "text")
```
