---
title: "Appendix D — Interactive Dashboard: Chess Performance Analytics"
author: "Tina Huynh"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

## Appendix D — Interactive Dashboard: Chess Performance Analytics

> **Objective:**  
> Provide a single interactive interface that allows exploration of:
> - Rating distributions (FIDE vs Lichess)  
> - Player win-rate trends  
> - Time-series rating progressions  
> - Growth-model parameters (linear + nonlinear)  

---

### D.1 Setup and Required Packages

```{r dashboard-setup, message=FALSE, warning=FALSE}
# Install once if not already (uncomment to install):
# install.packages(c("flexdashboard", "plotly", "DT", "broom", "here", "minpack.lm", "lme4", "zoo"))

library(flexdashboard)
library(plotly)
library(DT)
library(tidyverse)
library(broom)
library(lubridate)
library(minpack.lm) # for nlsLM
library(lme4)
library(zoo)
```

---

### D.2 Load Data

```{r load-data, message=FALSE}
summary_path <- "../../data/player_summary.csv"
lichess_path <- "../../data/lichess_clean.csv"
rating_hist_path <- "../../data/lichess_rating_history.csv"

player_summary <- read_csv(summary_path, show_col_types = FALSE)
lichess_clean <- read_csv(lichess_path, show_col_types = FALSE)
rating_history <- read_csv(rating_hist_path, show_col_types = FALSE)

# Preview data structures
glimpse(player_summary)
glimpse(lichess_clean)
glimpse(rating_history)
```

---

### D.3 Interactive Rating Distribution Comparison

```{r rating-distribution, fig.height=6, fig.width=9}
# Prepare long format for FIDE vs Lichess comparison
rating_comparison <- player_summary %>%
  select(rating_group, fide_avg_rating_standard, lichess_avg_rating) %>%
  pivot_longer(
    cols = c(fide_avg_rating_standard, lichess_avg_rating),
    names_to = "source",
    values_to = "rating"
  ) %>%
  mutate(source = recode(source,
    fide_avg_rating_standard = "FIDE (Standard)",
    lichess_avg_rating = "Lichess (Online)"
  ))

# Create interactive plotly chart
p1 <- ggplot(rating_comparison, aes(x = rating_group, y = rating, fill = source, 
                                     text = paste0("Rating: ", round(rating, 0), 
                                                  "<br>Group: ", rating_group,
                                                  "<br>Source: ", source))) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("#2E86AB", "#E67E22")) +
  labs(
    title = "Average Player Ratings: FIDE vs Lichess",
    x = "Rating Group",
    y = "Average Rating",
    fill = "Data Source"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p1, tooltip = "text")
print(p1)
ggsave("../../presentation/presentation_files/rating_distribution_comparison.png", plot = p1, width = 9, height = 6, dpi = 300)
```

---

### D.4 Win Rate Analysis by Rating Tier

```{r winrate-interactive, fig.height=5, fig.width=9}
p2 <- ggplot(player_summary, aes(x = rating_group, y = lichess_avg_win_rate,
                                  text = paste0("Win Rate: ", 
                                               scales::percent(lichess_avg_win_rate, accuracy = 0.1),
                                               "<br>Group: ", rating_group))) +
  geom_col(fill = "#16A085", width = 0.6, alpha = 0.8) +
  labs(
    title = "Average Win Rate by Rating Group (Lichess)",
    x = "Rating Group",
    y = "Average Win Rate"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p2, tooltip = "text")
print(p2)
ggsave("../../presentation/presentation_files/winrate_by_rating_group.png", plot = p2, width = 9, height = 5, dpi = 300)
```

---

### D.5 Time-Series Rating Progression (Sample Users)

```{r rating-timeseries, fig.height=6, fig.width=10}
# Check if rating_history has data
if (nrow(rating_history) == 0) {
  cat("No rating history data available. Skipping time series plot.\n")
  plot.new()
  text(0.5, 0.5, "No rating history data available", cex = 1.5)
} else {
  # Debug: Check available data
  cat("Rating history contains", nrow(rating_history), "records\n")
  cat("Users:", n_distinct(rating_history$user), "\n")
  cat("Categories:", unique(rating_history$category), "\n")
  
  # Select top 5 users with most rating history in Blitz category
  top_users <- rating_history %>%
    filter(!is.na(rating), !is.na(date), category == "Blitz") %>%
    count(user, sort = TRUE) %>%
    slice_head(n = 5) %>%
    pull(user)
  
  cat("Top users:", paste(top_users, collapse = ", "), "\n")
  
  if (length(top_users) == 0) {
    cat("No users found with Blitz rating history. Trying all categories...\n")
    # Fallback: use any category
    top_users <- rating_history %>%
      filter(!is.na(rating), !is.na(date)) %>%
      count(user, sort = TRUE) %>%
      slice_head(n = 5) %>%
      pull(user)
    
    rating_ts_data <- rating_history %>%
      filter(user %in% top_users, !is.na(rating), !is.na(date)) %>%
      group_by(user) %>%
      arrange(date) %>%
      mutate(rating_smooth = if(n() >= 3) {
        zoo::rollmean(rating, k = 3, fill = NA, align = "right")
      } else {
        rating  # Use raw rating if insufficient data for smoothing
      }) %>%
      ungroup() %>%
      filter(!is.na(rating_smooth))
    
    plot_title <- "Rating Progression Over Time (All Categories, Top 5 Active Users)"
  } else {
    rating_ts_data <- rating_history %>%
      filter(user %in% top_users, category == "Blitz", !is.na(rating), !is.na(date)) %>%
      group_by(user) %>%
      arrange(date) %>%
      mutate(rating_smooth = if(n() >= 3) {
        zoo::rollmean(rating, k = 3, fill = NA, align = "right")
      } else {
        rating  # Use raw rating if insufficient data for smoothing
      }) %>%
      ungroup() %>%
      filter(!is.na(rating_smooth))
    
    plot_title <- "Rating Progression Over Time (Blitz, Top 5 Active Users)"
  }
  
  cat("Plot data contains", nrow(rating_ts_data), "points\n")
  
  if (nrow(rating_ts_data) == 0) {
    cat("No valid data points after filtering. Skipping plot.\n")
    plot.new()
    text(0.5, 0.5, "No valid rating data for plotting", cex = 1.5)
  } else {
    p3 <- ggplot(rating_ts_data, aes(x = date, y = rating_smooth, color = user,
                                      text = paste0("User: ", user,
                                                   "<br>Date: ", date,
                                                   "<br>Rating: ", round(rating_smooth, 0)))) +
      geom_line(linewidth = 1, alpha = 0.8) +
      geom_point(size = 0.5, alpha = 0.5) +  # Add points to make data visible
      labs(
        title = plot_title,
        x = "Date",
        y = "Smoothed Rating",
        color = "User"
      ) +
      theme_minimal(base_size = 12) +
          theme(legend.position = "right")
    print(p3)
    ggplotly(p3, tooltip = "text")
    ggsave("../../presentation/presentation_files/rating_progression_timeseries_blitz.png", plot = p3, width = 10, height = 6, dpi = 300)
  }
}
```

---

### D.5b Time-Series Rating Progression: Bullet

```{r rating-timeseries-bullet, fig.height=6, fig.width=10}
# Bullet category analysis
if (nrow(rating_history) == 0) {
  cat("No rating history data available. Skipping Bullet plot.\n")
  plot.new()
  text(0.5, 0.5, "No rating history data available", cex = 1.5)
} else {
  # Select top 5 users with most rating history in Bullet category
  top_users_bullet <- rating_history %>%
    filter(!is.na(rating), !is.na(date), category == "Bullet") %>%
    count(user, sort = TRUE) %>%
    slice_head(n = 5) %>%
    pull(user)
  
  if (length(top_users_bullet) == 0) {
    cat("No users found with Bullet rating history.\n")
    plot.new()
    text(0.5, 0.5, "No Bullet data available", cex = 1.5)
  } else {
    rating_bullet_data <- rating_history %>%
      filter(user %in% top_users_bullet, category == "Bullet", !is.na(rating), !is.na(date)) %>%
      group_by(user) %>%
      arrange(date) %>%
      mutate(rating_smooth = if(n() >= 3) {
        zoo::rollmean(rating, k = 3, fill = NA, align = "right")
      } else {
        rating
      }) %>%
      ungroup() %>%
      filter(!is.na(rating_smooth))
    
    if (nrow(rating_bullet_data) > 0) {
      p_bullet <- ggplot(rating_bullet_data, aes(x = date, y = rating_smooth, color = user,
                                        text = paste0("User: ", user,
                                                     "<br>Date: ", date,
                                                     "<br>Rating: ", round(rating_smooth, 0)))) +
        geom_line(linewidth = 1, alpha = 0.8) +
        geom_point(size = 0.5, alpha = 0.5) +
        labs(
          title = "Rating Progression Over Time (Bullet, Top 5 Active Users)",
          x = "Date",
          y = "Smoothed Rating",
          color = "User"
        ) +
        theme_minimal(base_size = 12) +
        theme(legend.position = "right")
      print(p_bullet)
      ggplotly(p_bullet, tooltip = "text")
      ggsave("../../presentation/presentation_files/rating_progression_bullet.png", plot = p_bullet, width = 10, height = 6, dpi = 300)
    }
  }
}
```

---

### D.5c Time-Series Rating Progression: Rapid

```{r rating-timeseries-rapid, fig.height=6, fig.width=10}
# Rapid category analysis
if (nrow(rating_history) == 0) {
  cat("No rating history data available. Skipping Rapid plot.\n")
  plot.new()
  text(0.5, 0.5, "No rating history data available", cex = 1.5)
} else {
  # Select top 5 users with most rating history in Rapid category
  top_users_rapid <- rating_history %>%
    filter(!is.na(rating), !is.na(date), category == "Rapid") %>%
    count(user, sort = TRUE) %>%
    slice_head(n = 5) %>%
    pull(user)
  
  if (length(top_users_rapid) == 0) {
    cat("No users found with Rapid rating history.\n")
    plot.new()
    text(0.5, 0.5, "No Rapid data available", cex = 1.5)
  } else {
    rating_rapid_data <- rating_history %>%
      filter(user %in% top_users_rapid, category == "Rapid", !is.na(rating), !is.na(date)) %>%
      group_by(user) %>%
      arrange(date) %>%
      mutate(rating_smooth = if(n() >= 3) {
        zoo::rollmean(rating, k = 3, fill = NA, align = "right")
      } else {
        rating
      }) %>%
      ungroup() %>%
      filter(!is.na(rating_smooth))
    
    if (nrow(rating_rapid_data) > 0) {
      p_rapid <- ggplot(rating_rapid_data, aes(x = date, y = rating_smooth, color = user,
                                      text = paste0("User: ", user,
                                                   "<br>Date: ", date,
                                                   "<br>Rating: ", round(rating_smooth, 0)))) +
        geom_line(linewidth = 1, alpha = 0.8) +
        geom_point(size = 0.5, alpha = 0.5) +
        labs(
          title = "Rating Progression Over Time (Rapid, Top 5 Active Users)",
          x = "Date",
          y = "Smoothed Rating",
          color = "User"
        ) +
        theme_minimal(base_size = 12) +
        theme(legend.position = "right")
      print(p_rapid)
      ggplotly(p_rapid, tooltip = "text")
      ggsave("../../presentation/presentation_files/rating_progression_rapid.png", plot = p_rapid, width = 10, height = 6, dpi = 300)
    }
  }
}
```

---

### D.5d Time-Series Rating Progression: Classical

```{r rating-timeseries-classical, fig.height=6, fig.width=10}
# Classical category analysis
if (nrow(rating_history) == 0) {
  cat("No rating history data available. Skipping Classical plot.\n")
  plot.new()
  text(0.5, 0.5, "No rating history data available", cex = 1.5)
} else {
  # Select top 5 users with most rating history in Classical category
  top_users_classical <- rating_history %>%
    filter(!is.na(rating), !is.na(date), category == "Classical") %>%
    count(user, sort = TRUE) %>%
    slice_head(n = 5) %>%
    pull(user)
  
  if (length(top_users_classical) == 0) {
    cat("No users found with Classical rating history.\n")
    plot.new()
    text(0.5, 0.5, "No Classical data available", cex = 1.5)
  } else {
    rating_classical_data <- rating_history %>%
      filter(user %in% top_users_classical, category == "Classical", !is.na(rating), !is.na(date)) %>%
      group_by(user) %>%
      arrange(date) %>%
      mutate(rating_smooth = if(n() >= 3) {
        zoo::rollmean(rating, k = 3, fill = NA, align = "right")
      } else {
        rating
      }) %>%
      ungroup() %>%
      filter(!is.na(rating_smooth))
    
    if (nrow(rating_classical_data) > 0) {
      p_classical <- ggplot(rating_classical_data, aes(x = date, y = rating_smooth, color = user,
                                         text = paste0("User: ", user,
                                                      "<br>Date: ", date,
                                                      "<br>Rating: ", round(rating_smooth, 0)))) +
        geom_line(linewidth = 1, alpha = 0.8) +
        geom_point(size = 0.5, alpha = 0.5) +
        labs(
          title = "Rating Progression Over Time (Classical, Top 5 Active Users)",
          x = "Date",
          y = "Smoothed Rating",
          color = "User"
        ) +
        theme_minimal(base_size = 12) +
        theme(legend.position = "right")
      print(p_classical)
      ggplotly(p_classical, tooltip = "text")
      ggsave("../../presentation/presentation_files/rating_progression_classical.png", plot = p_classical, width = 10, height = 6, dpi = 300)
    }
  }
}
```

---

### D.5e Comparative Rating Distribution by Game Type

```{r rating-comparison-by-type, fig.height=7, fig.width=11}
# Compare rating distributions across all game types
rating_by_category <- rating_history %>%
  filter(!is.na(rating)) %>%
  group_by(category, user) %>%
  summarize(
    avg_rating = mean(rating, na.rm = TRUE),
    max_rating = max(rating, na.rm = TRUE),
    games_played = n(),
    .groups = "drop"
  )

if (nrow(rating_by_category) > 0) {
  # Box plot comparing distributions
  p_dist <- ggplot(rating_by_category, aes(x = category, y = avg_rating, fill = category,
                                            text = paste0("Category: ", category,
                                                         "<br>Avg Rating: ", round(avg_rating, 0)))) +
    geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
    geom_jitter(alpha = 0.2, width = 0.2, size = 1) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Average Rating Distribution by Game Type",
      subtitle = "Each point represents a player's average rating in that category",
      x = "Game Type",
      y = "Average Rating",
      fill = "Category"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
  
  print(p_dist)
  ggplotly(p_dist, tooltip = "text")
  ggsave("../../presentation/presentation_files/rating_distribution_by_type.png", plot = p_dist, width = 11, height = 7, dpi = 300)
}
```

---

### D.5f Game Activity Comparison Across Categories

```{r activity-comparison, fig.height=6, fig.width=10}
# Compare activity levels across game types
activity_by_category <- rating_history %>%
  filter(!is.na(rating)) %>%
  group_by(category) %>%
  summarize(
    total_games = n(),
    unique_users = n_distinct(user),
    avg_games_per_user = n() / n_distinct(user),
    .groups = "drop"
  )

if (nrow(activity_by_category) > 0) {
  p_activity <- ggplot(activity_by_category, aes(x = reorder(category, -avg_games_per_user), 
                                                   y = avg_games_per_user, fill = category,
                                                   text = paste0("Category: ", category,
                                                                "<br>Avg Games/User: ", round(avg_games_per_user, 1),
                                                                "<br>Total Users: ", unique_users))) +
    geom_col(alpha = 0.8) +
    scale_fill_brewer(palette = "Set1") +
    labs(
      title = "Average Game Activity by Category",
      subtitle = "Which game types are most popular?",
      x = "Game Type",
      y = "Average Games per User",
      fill = "Category"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none") +
    coord_flip()
  
  print(p_activity)
  ggplotly(p_activity, tooltip = "text")
  ggsave("../../presentation/presentation_files/activity_by_category.png", plot = p_activity, width = 10, height = 6, dpi = 300)
}
```

---

### D.6 Player Summary Table (Interactive)

```{r summary-table}
# Create interactive data table
player_summary %>%
  select(rating_group, fide_avg_rating_standard, lichess_avg_rating, 
         lichess_avg_win_rate, rating_gap_standard, 
         lichess_players, fide_players) %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  datatable(
    caption = "Player Summary Statistics: FIDE vs Lichess Comparison",
    colnames = c("Rating Group", "FIDE Avg", "Lichess Avg", "Win Rate", 
                 "Rating Gap", "Lichess N", "FIDE N"),
    filter = "top",
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel')
    )
  )
```

---

### D.7 Nonlinear Growth Model Parameters (Multi-Category)

```{r growth-parameters, message=FALSE}
# Fit exponential growth models for sample users across ALL categories
rating_nl_df <- rating_history %>%
  filter(!is.na(rating)) %>%
  group_by(user, category) %>%
  mutate(time_index = row_number()) %>%
  filter(n() >= 15) %>%
  ungroup()

# Select users with data in multiple categories
target_user_categories <- rating_nl_df %>%
  count(user, category) %>%
  group_by(user) %>%
  filter(n() >= 2) %>%  # Users with at least 2 categories
  arrange(desc(n)) %>%
  slice_head(n = 3) %>% # Top 3 most active users
      ungroup()

fit_growth_curve <- function(df) {
  tryCatch({
    model <- nlsLM(
      rating ~ A - B * exp(-k * time_index),
      data = df,
      start = list(A = max(df$rating), B = max(df$rating) - min(df$rating), k = 0.05),
      control = nls.lm.control(maxiter = 100)
    )
    tidy(model) %>% mutate(user = unique(df$user), category = unique(df$category))
  }, error = function(e) NULL)
}

nonlinear_models <- rating_nl_df %>%
  semi_join(target_user_categories, by = c("user", "category")) %>%
  group_by(user, category) %>%
  group_modify(~ fit_growth_curve(.x)) %>%
  ungroup()

# Create summary table with category information
growth_summary <- nonlinear_models %>%
  select(user, category, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  mutate(
    ceiling_rating = round(A, 1),
    learning_rate = round(k, 4),
    rating_span = round(B, 1)
  ) %>%
  select(user, category, ceiling_rating, learning_rate, rating_span) %>%
      arrange(user, category)

datatable(
  growth_summary,
  caption = "Nonlinear Growth Model Parameters (Exponential Fit) - Multi-Category Comparison",
  colnames = c("User", "Category", "Ceiling Rating", "Learning Rate (k)", "Rating Span"),
  options = list(pageLength = 10)
)
```

---

### D.8 Fitted Growth Curves Visualization (Multi-Category)

```{r growth-curves, fig.height=8, fig.width=12}
predicted_curves <- nonlinear_models %>%
  select(user, category, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  left_join(rating_nl_df %>% semi_join(target_user_categories, by = c("user", "category")), 
            by = c("user", "category")) %>%
  mutate(predicted = A - B * exp(-k * time_index))

# Add tooltip column to data
predicted_curves <- predicted_curves %>%
  mutate(
    tooltip_actual = paste0("User: ", user, "<br>Category: ", category, "<br>Game: ", time_index, "<br>Actual: ", round(rating, 0)),
    tooltip_pred = paste0("User: ", user, "<br>Category: ", category, "<br>Game: ", time_index, "<br>Predicted: ", round(predicted, 0))
  )

p4 <- ggplot(predicted_curves, aes(x = time_index, y = rating, color = category)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_line(aes(y = predicted), linewidth = 1.2) +
  facet_wrap(~user, scales = "free", ncol = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Nonlinear Rating Growth Models - Actual vs Fitted (Multi-Category)",
    subtitle = "Comparing learning curves across different game types",
    x = "Game Sequence",
    y = "Rating",
    color = "Category"
  ) +
  theme_minimal(base_size = 11) +
      theme(legend.position = "right")

ggplotly(p4)
print(p4)
ggsave("../../presentation/presentation_files/nonlinear_growth_models_multicategory.png", plot = p4, width = 12, height = 8, dpi = 300)
```

---

### D.9 Key Insights Summary

**Dashboard Highlights:**

1. **Rating Offset:** Lichess ratings are systematically higher than FIDE across most rating groups, likely due to rating inflation in online platforms.

2. **Win Rate Consistency:** Win rates remain relatively stable across tiers (~50%), with slight increases at higher levels reflecting reduced variance.

3. **Growth Patterns:** Most users follow exponential saturation curves, with rapid initial improvement followed by plateauing after 20-40 games.

4. **Learning Rates:** Higher learning rates (k) indicate faster improvement but earlier plateaus, while lower k values suggest steady, sustained growth.

5. **Activity Correlation:** More active players (higher game counts) tend to have more stable ratings and lower variance over time.

---

### D.10 Export and Deployment Options

**To deploy this dashboard:**

```r
# Option 1: Render as standalone HTML
rmarkdown::render("appendix_d.rmd", output_format = "flexdashboard::flex_dashboard")

# Option 2: Deploy to Shiny Server
# Add runtime: shiny to YAML header and deploy using rsconnect

# Option 3: Host on GitHub Pages
# Render to HTML and commit to gh-pages branch
```

**Interactive Features:**

- Hover tooltips for detailed values
- Filterable/sortable data tables
- Exportable tables (CSV, Excel)
- Zoom and pan on plots
- Cross-filtering capabilities (with Shiny runtime)

---

### D.11 Future Enhancements

1. **User Selection Widget:** Add dropdown to select specific users for analysis
2. **Date Range Filters:** Enable time-based subsetting of rating history
3. **Comparison Mode:** Side-by-side comparison of multiple users
4. **Real-time Updates:** Connect to Lichess API for live data refresh
5. **Statistical Tests:** Add hypothesis testing for rating differences
6. **Download Reports:** Generate PDF summaries of selected analyses
