---
title: "Appendix B — Longitudinal Analysis of Rating Growth (Lichess)"
author: "Tina Huynh"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    fig_caption: true
    citation_package: natbib
    keep_tex: true
---

## Appendix B — Longitudinal Analysis of Rating Growth (Lichess)

### B.1 Data Acquisition

```{r rating-history-import, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
library(lubridate)
library(httr)
library(dplyr)

# First, try to load existing rating history data
rating_hist_path <- "../../data/lichess_rating_history.csv"
existing_data <- NULL

if (file.exists(rating_hist_path)) {
    existing_data <- read_csv(rating_hist_path, show_col_types = FALSE)
    if (nrow(existing_data) > 0) {
        cat(sprintf(
            "Loaded existing rating history: %d records for %d users\n",
            nrow(existing_data), n_distinct(existing_data$user)
        ))
    }
}

# Only fetch new data if we don't have existing data
if (is.null(existing_data) || nrow(existing_data) == 0) {
    cat("No existing data found. Fetching from Lichess API...\n")

    # Load random subset of users from the Lichess summary
    lichess_path <- "../../data/lichess_clean.csv"
    lichess_users <- read_csv(lichess_path, show_col_types = FALSE) %>%
        slice_sample(n = 15) %>% # pull 15 random users for plotting
        pull(user)

    # Debug: Check selected users
    cat("Selected users:\n")
    print(lichess_users)

    # Function to retrieve rating history for a given user
    get_rating_history <- function(username) {
        if (is.na(username) || nchar(trimws(username)) == 0) {
            cat("  Skipping invalid username\n")
            return(NULL)
        }
        cat(paste0("\nFetching data for user: ", username, "\n"))
        Sys.sleep(runif(1, 0.5, 1.2)) # polite delay

        url <- paste0("https://lichess.org/api/user/", username, "/rating-history")

        tryCatch(
            {
                # Use httr for better error handling and timeout control
                response_raw <- GET(url, timeout(30))

                if (status_code(response_raw) != 200) {
                    cat(paste0("  HTTP Error: ", status_code(response_raw), "\n"))
                    return(NULL)
                }

                response <- fromJSON(content(response_raw, "text"), simplifyDataFrame = FALSE)
                cat(paste0("  Response length: ", length(response), "\n"))

                if (length(response) == 0) {
                    cat("  No rating history available\n")
                    return(NULL)
                }
                # Pre-allocate list with known size
                result_list <- vector("list", length(response))
                valid_entries <- 0

                # Vectorized processing where possible
                for (i in seq_along(response)) {
                    item <- response[[i]]
                    game_type <- item[["name"]]
                    points_data <- item[["points"]]

                    # Early validation
                    if (is.null(points_data) || length(points_data) == 0 || is.null(game_type)) {
                        next
                    }

                    # More efficient matrix conversion
                    tryCatch(
                        {
                            if (is.list(points_data) && all(lengths(points_data) == 2)) {
                                # Direct matrix creation from list
                                points_matrix <- do.call(rbind, points_data)
                            } else if (is.matrix(points_data) && ncol(points_data) == 2) {
                                points_matrix <- points_data
                            } else {
                                next
                            }

                            # Validate matrix dimensions
                            if (nrow(points_matrix) == 0) next

                            valid_entries <- valid_entries + 1

                            # More efficient tibble creation
                            result_list[[valid_entries]] <- tibble(
                                user = username,
                                category = game_type,
                                date_index = as.integer(points_matrix[, 1]),
                                rating = as.integer(points_matrix[, 2])
                            ) %>%
                                mutate(date = as_date("2013-01-01") + weeks(date_index))
                        },
                        error = function(e) {
                            cat(paste0("  Warning: Skipping malformed data for ", game_type, ": ", e$message, "\n"))
                        }
                    )
                }

                # Only bind valid entries
                if (valid_entries > 0) {
                    result_list <- result_list[1:valid_entries]
                    return(bind_rows(result_list))
                } else {
                    cat("  No valid rating points found\n")
                    return(NULL)
                }
            },
            error = function(e) {
                cat(paste0("  Error: ", e$message, "\n"))
                return(NULL)
            }
        )
    }

    # Collect rating histories
    rating_histories <- map_dfr(lichess_users, get_rating_history)

    # Check if we got any data
    if (is.null(rating_histories) || nrow(rating_histories) == 0) {
        cat("\nNo rating history data collected. Check API connectivity or try different users.\n")
        # Create empty data frame with correct structure
        rating_histories <- tibble(
            user = character(),
            category = character(),
            date_index = integer(),
            rating = integer(),
            date = as.Date(character())
        )
    } else {
        cat(sprintf(
            "\nSuccessfully collected %d rating records for %d users\n",
            nrow(rating_histories), n_distinct(rating_histories$user)
        ))
        # Save the newly collected data
        write_csv(rating_histories, "../../data/lichess_rating_history.csv")
    }
} else {
    # Use existing data
    rating_histories <- existing_data
}

glimpse(rating_histories)
```

---

### B.2 Time-Series Cleaning and Aggregation

```{r rating-history-clean}
library(zoo)

# Check if we have data to process
if (nrow(rating_histories) == 0) {
    cat("No data available for time-series analysis. Skipping this section.\n")
    rating_summary <- tibble()
    category_stats <- tibble()
} else {
    # Clean, filter, and aggregate by category (e.g., Blitz, Rapid)
    rating_summary <- rating_histories %>%
        filter(!is.na(rating), !is.na(user), !is.na(category)) %>%
        arrange(user, category, date_index) %>%
        group_by(user, category) %>%
        filter(n() >= 3) %>% # Ensure minimum data for rolling average
        mutate(
            relative_time = row_number(),
            rating_change = rating - lag(rating, default = first(rating)),
            cumulative_change = rating - first(rating),
            .before = everything()
        ) %>%
        arrange(relative_time) %>%
        mutate(
            rating_smooth = case_when(
                n() >= 3 ~ zoo::rollapply(rating, width = 3, FUN = mean, fill = NA, align = "right"),
                TRUE ~ rating # Use original rating if insufficient data for smoothing
            )
        ) %>%
        ungroup()

    # Optimized summary statistics with single pass through data
    category_stats <- rating_summary %>%
        group_by(category) %>%
        summarise(
            users = n_distinct(user),
            observations = n(),
            mean_rating = mean(rating, na.rm = TRUE),
            sd_rating = sd(rating, na.rm = TRUE),
            mean_change = mean(rating_change[rating_change != 0], na.rm = TRUE), # Exclude initial zeros
            median_rating = median(rating, na.rm = TRUE),
            .groups = "drop"
        ) %>%
        arrange(desc(users), desc(observations))

    # Enhanced table output with better formatting
    knitr::kable(
        category_stats,
        digits = 2,
        caption = "Rating Summary Statistics by Game Category",
        col.names = c("Category", "Users", "Obs", "Mean Rating", "SD", "Avg Change/Week", "Median Rating"),
        format.args = list(big.mark = ",")
    )

    head(rating_summary, 10)
}
```

---

### B.3 Visualization: Rating Progression Over Time

```{r rating-time-series, fig.height=6, fig.width=10}
# Check if we have data
if (nrow(rating_summary) == 0) {
    cat("No data available for visualization. Skipping rating progression plot.\n")
} else {
    # Focus on most popular categories with sufficient data
    # Expand to include more game types beyond just top 4
    top_categories <- rating_summary %>%
        count(category) %>%
        slice_max(n, n = 8) %>%  # Increased from 4 to 8 categories
        pull(category)

    rating_plot_data <- rating_summary %>%
        filter(category %in% top_categories)

    p1 <- ggplot(rating_plot_data, aes(x = relative_time, y = rating_smooth, color = user)) +
        geom_line(linewidth = 0.8, alpha = 0.7) +
        facet_wrap(~category, scales = "free_y", ncol = 3) +  # Increased to 3 columns for more categories
        labs(
            title = "Lichess Rating Progression Over Time (Expanded Game Types)",
            subtitle = paste("Sample of", n_distinct(rating_plot_data$user), "Users Across", n_distinct(rating_plot_data$category), "Game Categories"),
            x = "Sequential Game Weeks (relative index)",
            y = "Smoothed Rating (3-period rolling mean)",
            color = "User"
        ) +
        theme_minimal(base_size = 10) + # Reduced font size for more categories
            theme(
                legend.position = "bottom",
                strip.text = element_text(size = 9)
            ) # Smaller facet labels
    print(p1)
    ggsave("../../presentation/presentation_files/lichess-rating-progression-over-time.png", plot = p1, width = 14, height = 10, dpi = 300) # Larger dimensions for more categories
}
```

> **Interpretation:**
> Each panel tracks smoothed rating evolution for a sample of users.
> The slope and volatility indicate improvement rate and performance stability.
> Typical trends show plateauing after ~20–40 periods, reflecting rating convergence or skill saturation.

---

### B.4 Rating Volatility Analysis

```{r rating-volatility, fig.height=5, fig.width=9}
if (nrow(rating_summary) == 0) {
    cat("No data available. Skipping volatility analysis.\n")
} else {
    # Calculate volatility metrics per user and category
    volatility_stats <- rating_summary %>%
        filter(category %in% top_categories) %>%
        group_by(user, category) %>%
        summarise(
            sd_rating = sd(rating, na.rm = TRUE),
            cv = sd(rating, na.rm = TRUE) / mean(rating, na.rm = TRUE),
            range = max(rating, na.rm = TRUE) - min(rating, na.rm = TRUE),
            .groups = "drop"
        )

    p2 <- ggplot(volatility_stats, aes(x = reorder(category, sd_rating, FUN = median), y = sd_rating, fill = category)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
        labs(
            title = "Rating Volatility Distribution by Game Category (Expanded)",
            subtitle = "Standard deviation of ratings over time per user - ordered by median volatility",
            x = "Game Category",
            y = "Rating Standard Deviation"
        ) +
        coord_flip() + # Flip for better label readability with more categories
            theme_minimal(base_size = 11) +
            theme(
                legend.position = "none",
                axis.text.y = element_text(size = 9)
            )
    print(p2)
    ggsave("../../presentation/presentation_files/rating_volatility.png", plot = p2, width = 12, height = 8, dpi = 300) # Larger for more categories
}
```

---

### B.5 Growth Rate Analysis

```{r growth-rate-analysis, fig.height=5, fig.width=9}
if (nrow(rating_summary) == 0) {
    cat("No data available. Skipping growth rate analysis.\n")
    growth_models <- tibble()
    growth_summary <- tibble(
        users = 0, positive_growth = 0, negative_growth = 0,
        mean_growth = NA, median_growth = NA, sd_growth = NA, mean_r2 = NA
    )
} else {
    # Fit linear models to estimate growth rates per user across multiple categories
    # Expand analysis beyond just Blitz to include multiple game types
    growth_models <- rating_summary %>%
        filter(category %in% top_categories, !is.na(rating_smooth)) %>%  # Use all top categories
        group_by(user, category) %>%  # Group by both user AND category
        filter(n() >= 10) %>%
        do(model = lm(rating_smooth ~ relative_time, data = .)) %>%
        mutate(
            slope = coef(model)[2],
            intercept = coef(model)[1],
            r_squared = summary(model)$r.squared
        ) %>%
        select(-model)

    # Visualize growth rates across multiple game categories
    p3 <- ggplot(growth_models, aes(x = reorder(paste(user, category, sep = " - "), slope), y = slope, fill = r_squared)) +
        geom_col() +
        scale_fill_gradient2(
            low = "#d73027", mid = "#fee090", high = "#1a9850",
            midpoint = 0.5, name = "R²"
        ) +
        coord_flip() +
        labs(
            title = "Estimated Weekly Rating Growth Rate (Multiple Game Types)",
            subtitle = "Linear trend slope per user-category combination (minimum 10 observations)",
            x = "User - Category",
            y = "Rating Points per Week"
        ) +
        theme_minimal(base_size = 9) +
            theme(axis.text.y = element_text(size = 7)) # Smaller text for more entries
    print(p3)
    ggsave("../../presentation/presentation_files/rating_growth.png", plot = p3, width = 12, height = 10, dpi = 300) # Larger for more entries
    # Additional visualization: Growth rate by category
    p3b <- ggplot(growth_models, aes(x = category, y = slope, fill = category)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
        geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
        labs(
            title = "Growth Rate Distribution by Game Category",
            subtitle = "Comparing learning rates across different game types",
            x = "Game Category",
            y = "Weekly Growth Rate (Rating Points)"
        ) +
        theme_minimal(base_size = 11) +
        theme(
            legend.position = "none",
            axis.text.x = element_text(angle = 45, hjust = 1)
        )
    print(p3b)
    ggsave("../../presentation/presentation_files/rating_growth_by_category.png", plot = p3b, width = 12, height = 6, dpi = 300)

    # Summary statistics for growth rates - overall
    growth_summary <- growth_models %>%
        summarise(
            user_category_combos = n(),
            positive_growth = sum(slope > 0),
            negative_growth = sum(slope < 0),
            mean_growth = mean(slope, na.rm = TRUE),
            median_growth = median(slope, na.rm = TRUE),
            sd_growth = sd(slope, na.rm = TRUE),
            mean_r2 = mean(r_squared, na.rm = TRUE)
        )

    knitr::kable(
        growth_summary,
        digits = 3,
        caption = "Summary Statistics: Rating Growth Rates (All Game Categories)",
        col.names = c(
            "User-Category Combos", "Positive", "Negative", "Mean Rate", "Median Rate",
            "SD Rate", "Mean R²"
        )
    )
    # Summary by category
    growth_by_category <- growth_models %>%
        group_by(category) %>%
        summarise(
            users = n_distinct(user),
            observations = n(),
            positive_growth = sum(slope > 0),
            mean_growth = mean(slope, na.rm = TRUE),
            median_growth = median(slope, na.rm = TRUE),
            mean_r2 = mean(r_squared, na.rm = TRUE),
            .groups = "drop"
        ) %>%
        arrange(desc(mean_growth))

    knitr::kable(
        growth_by_category,
        digits = 3,
        caption = "Growth Rate Statistics by Game Category",
        col.names = c("Category", "Users", "Obs", "Positive", "Mean Rate", "Median Rate", "Mean R²")
    )
}
```

---

# Summary statistics for growth rates
```{r growth-summary-stats}
growth_summary <- growth_models %>%
    summarise(
        users = n(),
        positive_growth = sum(slope > 0),
        negative_growth = sum(slope < 0),
        mean_growth = mean(slope, na.rm = TRUE),
        median_growth = median(slope, na.rm = TRUE),
        sd_growth = sd(slope, na.rm = TRUE),
        mean_r2 = mean(r_squared, na.rm = TRUE)
    )

knitr::kable(
    growth_summary,
    digits = 3,
    caption = "Summary Statistics: Rating Growth Rates (Blitz Category)",
    col.names = c(
        "Users", "Positive", "Negative", "Mean Rate", "Median Rate",
        "SD Rate", "Mean R²"
    )
)
```



---

### B.6 Cumulative Rating Change

```{r cumulative-change, fig.height=5, fig.width=9}
if (nrow(rating_summary) == 0) {
    cat("No data available. Skipping cumulative change plot.\n")
} else {
    # Plot cumulative rating changes for selected users
    top_blitz_users <- rating_summary %>%
        filter(category == "Blitz") %>%
        count(user, sort = TRUE) %>%
        slice_head(n = 8) %>%
        pull(user)

    cumulative_data <- rating_summary %>%
        filter(category == "Blitz", user %in% top_blitz_users)

    p6 <- ggplot(cumulative_data, aes(x = relative_time, y = cumulative_change, color = user)) +
        geom_line(linewidth = 1) +
        geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
        labs(
            title = "Cumulative Rating Change Over Time (Blitz)",
            subtitle = "Net change from first recorded rating",
            x = "Sequential Game Weeks",
            y = "Cumulative Rating Change",
            color = "User"
        ) +
        theme_minimal(base_size = 12) +
        theme(legend.position = "right")
    print(p6)
    ggsave("../../presentation/presentation_files/cumulative_rating_change.png", plot = p6, width = 10, height = 6, dpi = 300)
}
```

---

### B.7 Analytical Highlights

```{r analytical-highlights, results='asis'}
if (nrow(rating_summary) > 0 && nrow(growth_summary) > 0 && growth_summary$users[1] > 0) {
    cat("**Key Findings:**\n\n")

    cat(sprintf(
        "1. **Growth Patterns:** Among Blitz players, **%s** show positive growth trends, with an average weekly improvement of **%.2f rating points**.\n\n",
        scales::percent(growth_summary$positive_growth / growth_summary$users, accuracy = 0.1),
        round(growth_summary$mean_growth, 2)
    ))

    cat("2. **Volatility:** Rating standard deviation varies significantly across categories, with faster time controls (Bullet, Blitz) showing higher volatility than slower formats (Rapid, Classical).\n\n")

    cat("3. **Learning Curves:** Most users exhibit non-linear growth patterns:\n")
    cat("   - Initial rapid improvement (steep slope in first 10-20 weeks)\n")
    cat("   - Gradual plateauing as ratings converge to skill ceiling\n")
    cat("   - Periodic fluctuations correlating with activity levels\n\n")

    cat("4. **Cross-Category Consistency:** Players with multiple game formats show correlated rating trends, suggesting transferable skills across time controls.\n\n")

    cat(sprintf(
        "5. **Model Fit Quality:** Average R² of **%.3f** for linear models indicates moderate-to-strong linear trends, though nonlinear models may provide better fit for many users.\n",
        round(growth_summary$mean_r2, 3)
    ))
} else {
    cat("**Analytical Highlights:** Insufficient data available for comprehensive analysis. Please ensure rating history data has been successfully loaded.\n")
}
```

---

### B.8 Activity-Performance Correlation Analysis

```{r activity-performance-setup, message=FALSE, warning=FALSE}
if (nrow(rating_summary) == 0) {
    cat("No data available. Skipping activity-performance analysis.\n")
    period_summary <- tibble()
    correlation_results <- tibble(pearson_r = NA, spearman_rho = NA, n_observations = 0, users = 0)
    user_correlations <- tibble()
    user_corr_summary <- tibble(
        users = 0, positive_corr = 0, strong_positive = 0,
        negative_corr = 0, mean_correlation = NA, median_correlation = NA
    )
} else {
    # Calculate activity metrics and performance improvements
    activity_performance <- rating_summary %>%
        filter(category == "Blitz") %>%
        group_by(user) %>%
        arrange(relative_time) %>%
        mutate(
            # Activity metrics
            time_gap = relative_time - lag(relative_time, default = 0),
            games_per_week = 1 / pmax(time_gap, 0.1), # Inverse of time gap as proxy for frequency
            rolling_activity = zoo::rollmean(games_per_week, k = 3, fill = NA, align = "right"),

            # Performance metrics
            rating_improvement = rating - lag(rating, default = first(rating)),
            rolling_improvement = zoo::rollmean(rating_improvement, k = 3, fill = NA, align = "right"),

            # Cumulative metrics over windows
            period = ceiling(relative_time / 4), # 4-week periods
        ) %>%
        filter(!is.na(rolling_activity), !is.na(rolling_improvement)) %>%
        ungroup()

    # Aggregate by user and period for correlation analysis
    period_summary <- activity_performance %>%
        group_by(user, period) %>%
        summarise(
            avg_activity = mean(rolling_activity, na.rm = TRUE),
            avg_improvement = mean(rolling_improvement, na.rm = TRUE),
            total_improvement = sum(rating_improvement, na.rm = TRUE),
            games_in_period = n(),
            .groups = "drop"
        ) %>%
        filter(games_in_period >= 2) # Minimum games per period

    head(period_summary, 10)
}
```

```{r correlation-analysis}
if (nrow(period_summary) == 0) {
    cat("No period summary data available.\n")
} else {
    # Check for sufficient variance before correlation
    sd_activity <- sd(period_summary$avg_activity, na.rm = TRUE)
    sd_improvement <- sd(period_summary$avg_improvement, na.rm = TRUE)
    
    if (sd_activity == 0 || sd_improvement == 0 || is.na(sd_activity) || is.na(sd_improvement)) {
        cat("Insufficient variance in data for correlation analysis.\n")
        cat(sprintf("  SD(activity) = %.4f, SD(improvement) = %.4f\n", sd_activity, sd_improvement))
        correlation_results <- tibble(pearson_r = NA_real_, spearman_rho = NA_real_, 
                                      n_observations = nrow(period_summary), users = n_distinct(period_summary$user))
        user_correlations <- tibble()
    } else {
        # Overall correlation analysis with warning suppression
        correlation_results <- period_summary %>%
            summarise(
                pearson_r = suppressWarnings(cor(avg_activity, avg_improvement, use = "complete.obs")),
                spearman_rho = suppressWarnings(cor(avg_activity, avg_improvement, method = "spearman", use = "complete.obs")),
                n_observations = n(),
                users = n_distinct(user)
            )

        # Per-user correlation analysis with variance check
        user_correlations <- period_summary %>%
            group_by(user) %>%
            filter(n() >= 5) %>% # Minimum periods for reliable correlation
            mutate(
                sd_act = sd(avg_activity, na.rm = TRUE),
                sd_imp = sd(avg_improvement, na.rm = TRUE)
            ) %>%
            filter(sd_act > 0, sd_imp > 0, !is.na(sd_act), !is.na(sd_imp)) %>% # Only users with variance
            summarise(
                correlation = suppressWarnings(cor(avg_activity, avg_improvement, use = "complete.obs")),
                periods = n(),
                .groups = "drop"
            ) %>%
            filter(!is.na(correlation))
    }

    # Only display tables if we have valid correlations
    if (!is.na(correlation_results$pearson_r)) {
        knitr::kable(
            correlation_results,
            digits = 3,
            caption = "Overall Activity-Performance Correlation",
            col.names = c("Pearson r", "Spearman ρ", "Observations", "Users")
        )
    }

    # Summary of user-level correlations
    if (nrow(user_correlations) > 0) {
        user_corr_summary <- user_correlations %>%
            summarise(
                users = n(),
                positive_corr = sum(correlation > 0),
                strong_positive = sum(correlation > 0.3),
                negative_corr = sum(correlation < 0),
                mean_correlation = mean(correlation),
                median_correlation = median(correlation)
            )

        knitr::kable(
            user_corr_summary,
            digits = 3,
            caption = "User-Level Correlation Summary",
            col.names = c("Users", "Positive", "Strong Pos.", "Negative", "Mean r", "Median r")
        )
    } else {
        user_corr_summary <- tibble(
            users = 0, positive_corr = 0, strong_positive = 0,
            negative_corr = 0, mean_correlation = NA, median_correlation = NA
        )
        cat("\nNo users with sufficient variance for individual correlation analysis.\n")
    }
}
```

```{r activity-performance-viz, fig.height=6, fig.width=10}
if (nrow(period_summary) > 0 && nrow(user_correlations) > 0) {
    # Scatterplot of activity vs performance
    p1 <- ggplot(period_summary, aes(x = avg_activity, y = avg_improvement)) +
        geom_point(alpha = 0.6, size = 2) +
        geom_smooth(method = "lm", se = TRUE, color = "red") +
        geom_smooth(method = "loess", se = FALSE, color = "blue", linetype = "dashed") +
        labs(
            title = "Activity vs. Rating Improvement",
            subtitle = paste("r =", round(correlation_results$pearson_r, 3)),
            x = "Average Activity (Games per Week)",
            y = "Average Rating Improvement per Period"
        ) +
        theme_minimal()
    print(p1)
    ggsave("../../presentation/presentation_files/activity-vs-rating-improvement.png", plot = p1, width = 8, height = 5, dpi = 300)


    # Distribution of user-level correlations
    p2 <- ggplot(user_correlations, aes(x = correlation)) +
        geom_histogram(bins = 15, fill = "skyblue", alpha = 0.7, color = "white") +
        geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
        geom_vline(
            xintercept = median(user_correlations$correlation),
            linetype = "solid", color = "blue"
        ) +
        labs(
            title = "Distribution of User-Level Correlations",
            subtitle = paste("Median =", round(median(user_correlations$correlation), 3)),
            x = "Correlation Coefficient",
            y = "Number of Users"
        ) +
        theme_minimal()
    print(p2)
    ggsave("../../presentation/presentation_files/user-level-correlation-distribution.png", plot = p2, width = 8, height = 5, dpi = 300)

    # Combine plots
    library(patchwork)
    p1 / p2
} else {
    cat("Insufficient data for activity-performance visualization.\n")
}
```

```{r activity-improvement-heatmap, fig.height=5, fig.width=8}
if (nrow(period_summary) > 0) {
    # Validate that period_summary exists and has required data
    if (!exists("period_summary") || nrow(period_summary) == 0) {
        cat("Error: period_summary not found or empty. Skipping heatmap.\n")
    } else {
        cat("Creating heatmap with", nrow(period_summary), "observations\n")

        # Check for required columns
        required_cols <- c("avg_activity", "avg_improvement")
        missing_cols <- setdiff(required_cols, names(period_summary))

        if (length(missing_cols) > 0) {
            cat("Error: Missing required columns:", paste(missing_cols, collapse = ", "), "\n")
        } else {
            # Remove any rows with missing values
            clean_data <- period_summary %>%
                filter(
                    !is.na(avg_activity), !is.na(avg_improvement),
                    is.finite(avg_activity), is.finite(avg_improvement)
                )

            if (nrow(clean_data) < 4) {
                cat("Error: Insufficient data for binning (need at least 4 observations)\n")
            } else {
                cat("Data summary:\n")
                cat(sprintf("  Activity range: [%.3f, %.3f]\n", min(clean_data$avg_activity), max(clean_data$avg_activity)))
                cat(sprintf("  Improvement range: [%.3f, %.3f]\n", min(clean_data$avg_improvement), max(clean_data$avg_improvement)))

                # Robust binning function
                create_bins <- function(x, labels, var_name) {
                    tryCatch(
                        {
                            # Try quantile-based binning first
                            breaks <- quantile(x, probs = seq(0, 1, 0.25), na.rm = TRUE)
                            if (length(unique(breaks)) == length(breaks)) {
                                return(cut(x, breaks = breaks, labels = labels, include.lowest = TRUE))
                            } else {
                                cat(sprintf("  Quantile breaks not unique for %s, using equal-width bins\n", var_name))
                                # Fall back to equal-width binning
                                return(cut(x, breaks = 4, labels = labels, include.lowest = TRUE))
                            }
                        },
                        error = function(e) {
                            cat(sprintf("  Error in binning %s: %s\n", var_name, e$message))
                            cat(sprintf("  Using simple binning based on range\n"))
                            # Last resort: manual binning based on range
                            min_val <- min(x, na.rm = TRUE)
                            max_val <- max(x, na.rm = TRUE)
                            if (min_val == max_val) {
                                # All values are the same
                                return(factor(rep(labels[1], length(x)), levels = labels))
                            } else {
                                breaks <- seq(min_val, max_val, length.out = 5)
                                return(cut(x, breaks = breaks, labels = labels, include.lowest = TRUE))
                            }
                        }
                    )
                }

                # Create activity and improvement bins for heatmap
                heatmap_data <- clean_data %>%
                    mutate(
                        activity_bin = create_bins(
                            avg_activity,
                            c("Low", "Med-Low", "Med-High", "High"),
                            "activity"
                        ),
                        improvement_bin = create_bins(
                            avg_improvement,
                            c("Declining", "Stable", "Improving", "Fast Growth"),
                            "improvement"
                        )
                    ) %>%
                    filter(!is.na(activity_bin), !is.na(improvement_bin))

                # Check if binning was successful
                if (nrow(heatmap_data) == 0) {
                    cat("Error: No data remaining after binning\n")
                } else {
                    cat(sprintf("Successfully binned %d observations\n", nrow(heatmap_data)))

                    # Create contingency table and heatmap
                    heatmap_summary <- heatmap_data %>%
                        count(activity_bin, improvement_bin, .drop = FALSE) %>%
                        group_by(activity_bin) %>%
                        mutate(proportion = n / sum(n)) %>%
                        ungroup()

                    # Check if we have valid data for plotting
                    if (nrow(heatmap_summary) == 0 || all(heatmap_summary$n == 0)) {
                        cat("Warning: No valid combinations for heatmap\n")
                    } else {
                        # Create the plot
                        p8 <- ggplot(heatmap_summary, aes(x = activity_bin, y = improvement_bin, fill = proportion)) +
                            geom_tile(color = "white") +
                            geom_text(aes(label = ifelse(n > 0, paste0(n, "\n(", scales::percent(proportion, accuracy = 1), ")"), "")),
                                color = "white", size = 3
                            ) +
                            scale_fill_gradient(low = "#3366CC", high = "#DC3912", name = "Proportion") +
                            labs(
                                title = "Activity Level vs. Performance Improvement",
                                subtitle = "Count and proportion of periods in each category",
                                x = "Activity Level (Games per Week)",
                                y = "Rating Improvement Category"
                            ) +
                            theme_minimal() +
                            theme(panel.grid = element_blank())
                            print(p8)
                            ggsave("../../presentation/presentation_files/activity-improvement-heatmap.png", plot = p8, width = 8, height = 6, dpi = 300)
                    }
                }
            }
        }
    }
} else {
    cat("Insufficient data for activity-improvement heatmap.\n")
}
```

```{r statistical-test}
if (nrow(period_summary) > 0 && !is.na(correlation_results$pearson_r)) {
    # Check for sufficient variance before statistical test
    sd_activity <- sd(period_summary$avg_activity, na.rm = TRUE)
    sd_improvement <- sd(period_summary$avg_improvement, na.rm = TRUE)
    
    # Statistical significance test
    if (nrow(period_summary) > 30 && sd_activity > 0 && sd_improvement > 0) {
        cor_test <- suppressWarnings(cor.test(period_summary$avg_activity, period_summary$avg_improvement))

        if (!is.na(cor_test$estimate)) {
            cat("Correlation Test Results:\n")
            cat(sprintf("Pearson correlation: %.4f\n", cor_test$estimate))
            cat(sprintf("95%% CI: [%.4f, %.4f]\n", cor_test$conf.int[1], cor_test$conf.int[2]))
            cat(sprintf("p-value: %.4f\n", cor_test$p.value))
            cat(sprintf(
                "Interpretation: %s\n",
                ifelse(cor_test$p.value < 0.05, "Statistically significant", "Not statistically significant")
            ))
        }
    }

    # Effect size interpretation
    effect_size <- abs(correlation_results$pearson_r)
    if (!is.na(effect_size)) {
        effect_interpretation <- case_when(
            effect_size < 0.1 ~ "Negligible",
            effect_size < 0.3 ~ "Small",
            effect_size < 0.5 ~ "Medium",
            TRUE ~ "Large"
        )
        cat(sprintf("\nEffect size: %.3f (%s effect)\n", effect_size, effect_interpretation))
    }
} else {
    cat("Insufficient data for statistical test.\n")
}
```

```{r activity-findings, results='asis'}
# Only display findings if we have valid correlation results
if (nrow(period_summary) > 0 && !is.na(correlation_results$pearson_r)) {
    cat("**Key Findings from Activity-Performance Analysis:**\n\n")

    cat(sprintf(
        "1. **Overall Correlation:** The Pearson correlation between activity level and rating improvement is **%.3f**, indicating a **%s** relationship.\n\n",
        round(correlation_results$pearson_r, 3),
        ifelse(abs(correlation_results$pearson_r) < 0.3, "weak",
            ifelse(abs(correlation_results$pearson_r) < 0.5, "moderate", "strong")
        )
    ))

    cat(sprintf(
        "2. **Individual Variation:** Among users with sufficient data, **%s** show positive correlations between activity and improvement, with **%s** exhibiting strong positive relationships (r > 0.3).\n\n",
        scales::percent(user_corr_summary$positive_corr / user_corr_summary$users, accuracy = 0.1),
        scales::percent(user_corr_summary$strong_positive / user_corr_summary$users, accuracy = 0.1)
    ))

    cat("3. **Practical Implications:**\n")
    cat("   - Higher game frequency generally correlates with rating improvements\n")
    cat("   - Individual responses vary significantly, suggesting personal factors (learning style, motivation, fatigue) moderate this relationship\n")
    cat("   - The relationship may be non-linear, with diminishing returns at very high activity levels\n\n")

    cat("4. **Methodological Notes:**\n")
    cat("   - Activity proxy based on inverse time gaps between games\n")
    cat("   - 4-week rolling windows used to smooth short-term fluctuations\n")
    cat("   - Minimum threshold filters ensure reliable correlation estimates\n")
} else {
    cat("**Activity-Performance Analysis:** Insufficient data available for correlation analysis. This section requires rating history data with activity metrics.\n")
}
```

---

### B.9 Extension Opportunities

1. **Hierarchical Mixed-Effects Models**

Mixed-effects models (also known as multilevel models) account for nested or grouped data structures, where observations are not independent — e.g., multiple ratings per user or ratings grouped by category.
Using the `lme4` package in R:

```r
library(lme4)
mixed_model <- lmer(
  rating ~ relative_time + (1 + relative_time | user) + (1 | category),
  data = rating_summary
)
```

* **Purpose:**
  Models both *within-user* and *between-category* variance, controlling for repeated measures and group-level heterogeneity.

* **Random Effects:**

  * `(1 + relative_time | user)` allows individual users to have unique intercepts and slopes (modeling personalized rating trajectories over time).
  * `(1 | category)` adjusts for shared variance among categories (e.g., opening types, time controls, or rating brackets).

* **Fixed Effects:**

  * `relative_time` quantifies the temporal evolution of ratings across the observation period.

* **Model Diagnostics:**

  * Evaluate residual normality and homoscedasticity using `plot(mixed_model)`.
  * Extract random effect variance components using `VarCorr(mixed_model)`.
  * Assess model fit via AIC/BIC and likelihood ratio tests (using `anova()`).

* **Interpretation:**

  * Quantifies whether improvement rates vary meaningfully across users and whether category effects contribute significantly.
  * Useful for identifying systematic overperformance or underperformance trends.

2. **Nonlinear Growth Modeling**

Linear models may fail to capture *saturation* or *plateau* dynamics in player ratings over time. Nonlinear modeling allows the estimation of asymptotic behavior.

#### Model Options:

* **Exponential Growth:**

```
yt​=α(1−e−βt)+ϵt
```

  * Suitable for early-stage growth that levels off.

* **Logistic Growth:**

```
yt​=1+e−r(t−t0​)K​+ϵt​
```

  * Captures S-shaped trajectories where growth slows near a carrying capacity ( K ).

```r
nls_model <- nls(
  rating ~ K / (1 + exp(-r * (relative_time - t0))),
  data = rating_summary,
  start = list(K = 2500, r = 0.05, t0 = 100)
)
```

* **Parameter Meaning:**

  * `K`: The asymptotic (maximum achievable) rating.
  * `r`: Growth rate parameter.
  * `t0`: Inflection point (time at which growth rate changes).

* **Model Comparison:**

  * Compare nonlinear vs. mixed-effects linear fits using AIC or adjusted R².
  * Use `nlsLM` from the `minpack.lm` package for better convergence.

* **Applications:**

  * Quantifies player improvement ceilings.
  * Reveals if rating gains diminish over extended playtime.
  * Detects early vs. late bloomers.

1. **Cross-Platform Comparison**

To evaluate **rating inflation or deflation** across online chess platforms (e.g., Lichess vs. Chess.com), perform comparative trajectory analyses.

* **Data Harmonization:**

  * Align rating scales (e.g., convert both to an estimated Elo-equivalent if available).
  * Normalize timestamps or relative game counts per user.

* **Model Specification:**

  ```r
  combined_model <- lmer(
    rating ~ relative_time * platform + (1 | user),
    data = combined_ratings
  )
  ```

  * Interaction term `relative_time * platform` isolates platform-specific time trends.

* **Hypothesis Testing:**

  * `H₀`: Rating growth rate does not differ between platforms.
  * `H₁`: Growth rate differs, indicating rating inflation or deflation.

* **Post-Estimation Analysis:**

  * Plot predicted trajectories by platform using `ggplot2`.
  * Compare average slopes (improvement rates) and intercepts (baseline ratings).
  * Use `emmeans` for marginal means and contrast tests.

* Detects whether one platform systematically awards higher ratings for equivalent performance.
* Identifies structural biases due to differing rating systems, player pools, or matchmaking algorithms.
