## Appendix A — Summary Analytics: Lichess vs. FIDE Rating Patterns

### A.1 Data Import and Preparation

```{r import-summary, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)

# Define correct paths relative to repo structure
summary_path <- "../data/player_summary.csv"

# Import the summary dataset generated by data_import.R
player_summary <- read_csv(summary_path, show_col_types = FALSE)

# Verify import and inspect structure
glimpse(player_summary)
summary(player_summary)
```

---

### A.2 Descriptive Overview

```{r descriptive-overview, message=FALSE}
# Summary statistics table
player_summary %>%
  select(rating_group, fide_avg_rating, lichess_avg_rating, avg_win_rate, performance_gap, lichess_players, fide_players) %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  knitr::kable(
    caption = "Summary of Mean Ratings, Win Rates, and Player Counts by Rating Group",
    col.names = c("Rating Group", "FIDE Avg Rating", "Lichess Avg Rating", "Avg Win Rate", "Performance Gap", "Lichess N", "FIDE N")
  )
```

> **Interpretation:**
> This table compares average rating levels and player volumes between the simulated FIDE dataset and randomly sampled Lichess players.
> The *Performance Gap* represents how much higher (or lower) the Lichess average ratings appear relative to the corresponding FIDE rating band.

---

### A.3 Visualization: Average Rating Comparison

```{r rating-comparison-plot, fig.height=4.5, fig.width=7}
player_summary_long <- player_summary %>%
  pivot_longer(
    cols = c(fide_avg_rating, lichess_avg_rating),
    names_to = "source", values_to = "rating"
  ) %>%
  mutate(source = recode(source,
    fide_avg_rating = "FIDE (Blitz)",
    lichess_avg_rating = "Lichess (Online)"
  ))

ggplot(player_summary_long, aes(x = rating_group, y = rating, fill = source)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = round(rating, 0)), position = position_dodge(width = 0.7), vjust = -0.5, size = 3.2) +
  scale_fill_manual(values = c("#2E86AB", "#E67E22")) +
  labs(
    title = "Average Player Ratings: FIDE vs. Lichess",
    subtitle = "Comparison by Rating Group (Simulated Distribution, n ≈ 500–900 each)",
    x = "Rating Group",
    y = "Average Rating",
    fill = "Data Source"
  ) +
  theme_minimal(base_size = 12)
```

> **Observation:**
> Lichess rating averages may differ systematically from FIDE values due to differing player pools and rating inflation between online and over-the-board ecosystems.
> The mid-tier groups typically show closer alignment, while extreme ends (Under-1600 and 2600+) exhibit greater divergence.

---

### A.4 Visualization: Win Rate and Game Volume by Tier

```{r winrate-games, fig.height=4.5, fig.width=7}
ggplot(player_summary, aes(x = rating_group, y = avg_win_rate)) +
  geom_col(fill = "#16A085", width = 0.6, alpha = 0.8) +
  geom_text(aes(label = percent(avg_win_rate, accuracy = 0.1)), vjust = -0.3, size = 3.5) +
  labs(
    title = "Average Win Rate Across Rating Groups (Lichess Sample)",
    subtitle = "Higher tiers generally exhibit increased consistency and lower variance",
    x = "Rating Group",
    y = "Average Win Rate"
  ) +
  theme_minimal(base_size = 12)
```

```{r games-volume, fig.height=4.5, fig.width=7}
ggplot(player_summary, aes(x = rating_group, y = avg_games)) +
  geom_point(size = 4, color = "#2980B9") +
  geom_line(aes(group = 1), color = "#2980B9", linewidth = 1) +
  labs(
    title = "Average Number of Games Played by Rating Tier (Lichess)",
    subtitle = "More active users show gradual rating stability and reduced volatility",
    x = "Rating Group",
    y = "Average Games Played"
  ) +
  theme_light(base_size = 12)
```

---

### A.5 Analytical Commentary

**Summary of Findings:**

1. There is a measurable **rating offset** between online (Lichess) and official (FIDE) distributions, reflecting the structural differences in rating systems.
2. **Win rates** are relatively stable across rating groups, with a mild upward trend among higher-rated users, consistent with greater consistency and lower variance.
3. The **number of games played** increases modestly with rating, implying experience and activity correlate with performance sustainability.
4. The **performance gap** metric provides a proxy for inflation or competitive bias — a useful baseline for future normalization across rating systems.
