## Appendix B — Longitudinal Analysis of Rating Growth (Lichess)

### B.1 Data Acquisition

```{r rating-history-import, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
library(lubridate)
library(httr)

# Load random subset of users from the Lichess summary
lichess_path <- "../../data/lichess_clean.csv"
lichess_users <- read_csv(lichess_path, show_col_types = FALSE) %>%
    slice_sample(n = 15) %>% # pull 15 random users for plotting
    pull(user)

# Debug: Check selected users
cat("Selected users:\n")
print(lichess_users)

# Function to retrieve rating history for a given user
get_rating_history <- function(username) {
    cat(paste0("\nFetching data for user: ", username, "\n"))
    Sys.sleep(runif(1, 0.5, 1.2)) # polite delay
    url <- paste0("https://lichess.org/api/user/", username, "/rating-history")
    tryCatch(
        {
            response <- fromJSON(url, simplifyDataFrame = FALSE)
            cat(paste0("  Response length: ", length(response), "\n"))
            if (length(response) == 0) {
                cat("  No rating history available\n")
                return(NULL)
            }
            
            # Process each game type
            result_list <- list()
            for (i in seq_along(response)) {
                item <- response[[i]]
                game_type <- item[["name"]]
                points_data <- item[["points"]]
                
                if (!is.null(points_data) && length(points_data) > 0) {
                    # Handle different possible structures
                    if (is.list(points_data) && !is.data.frame(points_data)) {
                        # Convert list of vectors to matrix
                        points_matrix <- matrix(unlist(points_data), ncol = 2, byrow = TRUE)
                    } else if (is.matrix(points_data)) {
                        points_matrix <- points_data
                    } else {
                        next  # Skip if unexpected format
                    }
                    
                    result_list[[i]] <- tibble(
                        user = username,
                        category = game_type,
                        date_index = points_matrix[, 1],
                        rating = points_matrix[, 2]
                    ) %>%
                        mutate(date = as_date("2013-01-01") + weeks(date_index))
                }
            }
            
            if (length(result_list) > 0) {
                return(bind_rows(result_list))
            } else {
                cat("  No valid rating points found\n")
                return(NULL)
            }
        },
        error = function(e) {
            cat(paste0("  Error: ", e$message, "\n"))
            return(NULL)
        }
    )
}

# Collect rating histories
rating_histories <- map_dfr(lichess_users, get_rating_history)
write_csv(rating_histories, "../../data/lichess_rating_history.csv")
glimpse(rating_histories)
```

---

### B.2 Time-Series Cleaning and Aggregation

```{r rating-history-clean}
# Clean, filter, and aggregate by category (e.g., Blitz, Rapid)
rating_summary <- rating_histories %>%
    filter(!is.na(rating)) %>%
    group_by(user, category) %>%
    mutate(
        rating_smooth = zoo::rollapply(rating, width = 3, FUN = mean, fill = NA, align = "right"),
        relative_time = row_number()
    ) %>%
    ungroup()

head(rating_summary)
```

---

### B.3 Visualization: Rating Progression Over Time

```{r rating-time-series, fig.height=5, fig.width=8}
ggplot(rating_summary, aes(x = relative_time, y = rating_smooth, color = user)) +
    geom_line(linewidth = 1) +
    facet_wrap(~category, scales = "free_y") +
    labs(
        title = "Lichess Rating Progression Over Time",
        subtitle = "Sample of 5 Randomly Selected Users Across Game Categories",
        x = "Sequential Game Weeks (relative index)",
        y = "Smoothed Rating (3-period rolling mean)",
        color = "User"
    ) +
    theme_minimal(base_size = 12)
```

> **Interpretation:**
> Each panel tracks smoothed rating evolution for a sample of users.
> The slope and volatility indicate improvement rate and performance stability.
> Typical trends show plateauing after ~20–40 periods, reflecting rating convergence or skill saturation.

---

### B.4 Analytical Highlights

* **Consistent upward trends** imply sustained learning and adaptation to faster time controls.
* **Volatile fluctuations** correspond to periods of experimentation or inactivity.
* Cross-category gaps (e.g., *Blitz > Rapid > Classical*) align with expected behavioral rating hierarchies in online chess.
* Aggregating trajectories could enable a **growth-rate model** (e.g., mixed-effects regression) quantifying individual improvement rates.

---

### B.5 Extension Opportunities

1. **Model Rating Growth:**
   Fit a linear or non-linear regression for each user to estimate weekly growth rate.

   ```r
   model <- lm(rating_smooth ~ relative_time, data = rating_summary)
   summary(model)
   ```

2. **Compare Variability by Rating Group:**
   Merge `rating_summary` with the existing `player_summary.csv` to test whether higher-rated players show less variance over time.

3. **Add Chess.com Parallel Data:**
   Use Chess.com’s public API to collect equivalent histories and benchmark cross-platform consistency.
