---
title: "Final Project Proposal - Statistical Analysis of Chess Performance and Player Progression"
author: "Tina Huynh"
date: "`r Sys.Date()`"
output: html_document
---

# Statistical Analysis of Chess Performance and Player Progression

## Information

### Research Questions

1. How do **opening choices** influence win probability across time controls (e.g., blitz, rapid, classical)?
2. Does **playing frequency** or **game volume** correlate with **rating growth** over time?
3. Are **white pieces** statistically more likely to win, and does this advantage vary by rating or game type?
4. How does **player rating** predict **accuracy and move efficiency**, and are these patterns consistent across skill brackets?

---

### Data Information

#### Lichess Public API and Game Database

The **Lichess API** is an open data service provided by Lichess.org, a free online chess platform.  
It offers access to detailed, anonymized game records, including metadata such as player ratings, time controls, openings, and results.  
The **Lichess Database** also publishes monthly bulk archives of millions of games in PGN format, which can be converted into CSV or JSON for analysis.
## Data Analysis Plan

### Variable and Function Naming Conventions

Variables will use lowercase snake_case naming (e.g., `rating_diff`, `win_rate`, `avg_move_time`).  
Custom functions will follow similar naming conventions, such as `calc_win_rate()` and `normalize_rating_change()`.  
Data will be stored in tibbles and manipulated with `dplyr` and `tidyr`.
---

### Data Cleaning

1. Import JSON or CSV data from the Lichess API archive.  
2. Standardize rating fields, convert timestamps, and filter out corrupted games.  
3. Recode categorical fields for uniformity (`winner`, `color`, `game_type`).  
4. Derive computed variables such as:
   - `rating_change = rating_post - rating_pre`
   - `is_white_win = ifelse(winner == "white", 1, 0)`
5. Integrate FIDE data by matching players or aggregating by country/title.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(httr)
library(jsonlite)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(skimr)
library(corrplot)
library(broom)
library(modelr)
library(scales)

# Set theme for consistent plotting
theme_set(theme_minimal())
```
---

### Comparison Groups

#### Primary Grouping Variables

1. **By Color:** White vs. Black
   - **First-move advantage analysis**: Quantifying the theoretical 52-56% win rate for white pieces
   - **Opening initiative impact**: How color affects opening choice and early game strategy
   - **Endgame conversion rates**: Color-specific success in winning/drawing positions
   - **Psychological factors**: Player confidence and aggression based on piece color

2. **By Game Type:** Blitz, Rapid, Classical, Bullet
   - **Bullet (< 3 minutes)**: Ultra-fast games emphasizing intuition and pattern recognition
   - **Blitz (3-8 minutes)**: Fast-paced games balancing speed and tactical calculation
   - **Rapid (8-25 minutes)**: Moderate time controls allowing strategic planning
   - **Classical (> 25 minutes)**: Deep positional games with extensive calculation
   - **Correspondence**: Turn-based games with days per move (if available)
#### Statistical Comparison Framework

**Pairwise Comparisons:**
- **Two-sample t-tests** for continuous variables (rating, accuracy)
- **Chi-square tests** for categorical outcomes (win/loss/draw)
- **Wilcoxon rank-sum tests** for non-parametric comparisons
- **Fisher's exact tests** for small sample contingency tables
#### Expected Group Differences

**Color-based Expectations:**
- White advantage should be consistent across rating levels (52-56% win rate)
- Advantage may diminish in bullet games due to time pressure
- Opening diversity should favor white pieces in aggressive systems
#### Control Variables and Confounding Factors

**Player-level Controls:**
- Historical rating trajectory (improving vs. declining)
- Playing frequency and recent activity
- Specialization in specific time controls or openings
- Title status and competitive experience

**Game-level Controls:**
- Opponent strength relative to player rating
- Recent performance streak effects
- Opening preparation depth
- Hardware/software advantages (engine assistance detection)

**Environmental Controls:**
- Platform-specific features (Lichess vs. Chess.com differences)
- Time zone and circadian rhythm effects
- Internet connection stability impacts
- Mobile vs. desktop playing interface

#### Sample Size and Power Analysis

**Minimum Sample Requirements:**
- **Color analysis**: 100+ games per color for reliable estimates
- **Time control comparison**: 50+ games per category
- **Rating tier analysis**: 30+ games per 100-point rating band
- **Opening effectiveness**: 20+ games per opening family

**Power Calculations:**
- **Effect size detection**: Minimum 5% win rate difference
- **Statistical power**: Target 80% power for primary comparisons
- **Alpha level**: 0.05 with Bonferroni correction for multiple tests
- **Confidence intervals**: 95% CI for all effect estimates
