---
title: "Final Project Proposal - Statistical Analysis of Chess Performance and Player Progression"
author: "Tina Huynh"
date: "`r Sys.Date()`"
output: html_document
---

# Statistical Analysis of Chess Performance and Player Progression

## Information

### Research Questions

1. How do **opening choices** influence win probability across time controls (e.g., blitz, rapid, classical)?
2. Does **playing frequency** or **game volume** correlate with **rating growth** over time?
3. Are **white pieces** statistically more likely to win, and does this advantage vary by rating or game type?
4. How does **player rating** predict **accuracy and move efficiency**, and are these patterns consistent across skill brackets?

---

### Data Information

#### Lichess Public API and Game Database

The **Lichess API** is an open data service provided by Lichess.org, a free online chess platform.  
It offers access to detailed, anonymized game records, including metadata such as player ratings, time controls, openings, and results.  
The **Lichess Database** also publishes monthly bulk archives of millions of games in PGN format, which can be converted into CSV or JSON for analysis.
## Data Analysis Plan

### Variable and Function Naming Conventions

Variables will use lowercase snake_case naming (e.g., `rating_diff`, `win_rate`, `avg_move_time`).  
Custom functions will follow similar naming conventions, such as `calc_win_rate()` and `normalize_rating_change()`.  
Data will be stored in tibbles and manipulated with `dplyr` and `tidyr`.
---

### Data Cleaning

1. Import JSON or CSV data from the Lichess API archive.  
2. Standardize rating fields, convert timestamps, and filter out corrupted games.  
3. Recode categorical fields for uniformity (`winner`, `color`, `game_type`).  
4. Derive computed variables such as:
   - `rating_change = rating_post - rating_pre`
   - `is_white_win = ifelse(winner == "white", 1, 0)`
5. Integrate FIDE data by matching players or aggregating by country/title.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(httr)
library(jsonlite)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(skimr)
library(corrplot)
library(broom)
library(modelr)
library(scales)

# Set theme for consistent plotting
theme_set(theme_minimal())
```

```{r data-collection}
# Enhanced data collection function with error handling and rate limiting
collect_lichess_data <- function(username, max_games = 500, delay = 1) {
  cat("Collecting data for user:", username, "\n")
  
  # API request with proper error handling
  url <- paste0("https://lichess.org/api/games/user/", username, 
                "?max=", max_games, "&analysed=true&perfType=blitz,rapid,classical")
  
  tryCatch({
    response <- GET(url, 
                   add_headers("Accept" = "application/x-ndjson"),
                   timeout(30))
    
    if (status_code(response) != 200) {
      stop("API request failed with status: ", status_code(response))
    }
    
    raw_text <- content(response, as = "text", encoding = "UTF-8")
    
    if (nchar(raw_text) == 0) {
      warning("Empty response from API")
      return(tibble())
    }
    
    # Split and parse NDJSON
    lines <- strsplit(raw_text, "\n")[[1]]
    lines <- lines[nzchar(lines)]
    
    cat("Processing", length(lines), "games...\n")
    
    # Enhanced parsing with better error handling
    lichess_games <- map(lines, function(line) {
      tryCatch({
        obj <- fromJSON(line, flatten = TRUE)
        
        # Extract player-specific data
        user_color <- if(!is.null(obj$players$white$user$id) && 
                        obj$players$white$user$id == username) "white" else "black"
        
        # More comprehensive data extraction
        tibble(
          id = obj$id %||% NA_character_,
          username = username,
          user_color = user_color,
          rated = obj$rated %||% NA,
          speed = obj$speed %||% NA_character_,
          perf = obj$perf %||% NA_character_,
          createdAt = obj$createdAt %||% NA_real_,
          lastMoveAt = obj$lastMoveAt %||% NA_real_,
          turns = obj$turns %||% NA_integer_,
          winner = obj$winner %||% NA_character_,
          status = obj$status %||% NA_character_,
          white_rating = obj$players$white$rating %||% NA_integer_,
          black_rating = obj$players$black$rating %||% NA_integer_,
          white_username = obj$players$white$user$id %||% NA_character_,
          black_username = obj$players$black$user$id %||% NA_character_,
          opening_name = obj$opening$name %||% NA_character_,
          opening_eco = obj$opening$eco %||% NA_character_,
          opening_ply = obj$opening$ply %||% NA_integer_,
          clock_initial = obj$clock$initial %||% NA_integer_,
          clock_increment = obj$clock$increment %||% NA_integer_,
          clock_total = obj$clock$totalTime %||% NA_integer_,
          # Analysis data if available
          white_accuracy = obj$players$white$analysis$accuracy %||% NA_real_,
          black_accuracy = obj$players$black$analysis$accuracy %||% NA_real_,
          white_blunder = obj$players$white$analysis$blunder %||% NA_integer_,
          black_blunder = obj$players$black$analysis$blunder %||% NA_integer_
        )
      }, error = function(e) {
        warning("Failed to parse line: ", substr(line, 1, 100), "...")
        return(NULL)
      })
    })
    
    # Compact and bind
    lichess_games <- compact(lichess_games)
    
    if (length(lichess_games) == 0) {
      warning("No valid games parsed")
      return(tibble())
    }
    
    result_df <- bind_rows(lichess_games)
    cat("Successfully collected", nrow(result_df), "games\n")
    
    # Rate limiting
    if (delay > 0) Sys.sleep(delay)
    
    return(result_df)
    
  }, error = function(e) {
    warning("Data collection failed: ", e$message)
    return(tibble())
  })
}
glimpse(lichess_df)
```
```

---

### Comparison Groups

#### Primary Grouping Variables

1. **By Color:** White vs. Black
   - **First-move advantage analysis**: Quantifying the theoretical 52-56% win rate for white pieces
   - **Opening initiative impact**: How color affects opening choice and early game strategy
   - **Endgame conversion rates**: Color-specific success in winning/drawing positions
   - **Psychological factors**: Player confidence and aggression based on piece color

2. **By Game Type:** Blitz, Rapid, Classical, Bullet
   - **Bullet (< 3 minutes)**: Ultra-fast games emphasizing intuition and pattern recognition
   - **Blitz (3-8 minutes)**: Fast-paced games balancing speed and tactical calculation
   - **Rapid (8-25 minutes)**: Moderate time controls allowing strategic planning
   - **Classical (> 25 minutes)**: Deep positional games with extensive calculation
   - **Correspondence**: Turn-based games with days per move (if available)

3. **By Rating Tier:**
   - **Beginner (< 1000)**: Learning basic rules and piece values
   - **Novice (1000-1400)**: Developing tactical awareness and opening principles
   - **Intermediate (1400–1800)**: Strategic understanding and pattern recognition
   - **Advanced (1800–2200)**: Deep positional play and endgame technique
   - **Expert (2200-2400)**: Near-master level with specialized knowledge
   - **Master (2400+)**: Elite players with comprehensive chess understanding

4. **By Opening Type:** 
   - **Aggressive Openings**: King's Gambit, Evans Gambit, Danish Gambit
   - **Positional Openings**: Ruy Lopez, Queen's Gambit, English Opening
   - **Hypermodern Systems**: King's Indian Defense, Nimzo-Indian Defense
   - **Solid Defenses**: French Defense, Caro-Kann Defense, Petrov Defense
   - **Sicilian Variations**: Open Sicilian, Closed Sicilian, Dragon Variation
#### Statistical Comparison Framework

**Pairwise Comparisons:**
- **Two-sample t-tests** for continuous variables (rating, accuracy)
- **Chi-square tests** for categorical outcomes (win/loss/draw)
- **Wilcoxon rank-sum tests** for non-parametric comparisons
- **Fisher's exact tests** for small sample contingency tables

**Multi-group Analysis:**
- **ANOVA** for comparing means across multiple groups
- **Kruskal-Wallis tests** for non-parametric multi-group comparisons
- **Post-hoc testing** with Bonferroni correction for multiple comparisons
- **Effect size calculations** (Cohen's d, eta-squared) for practical significance
#### Expected Group Differences

**Color-based Expectations:**
- White advantage should be consistent across rating levels (52-56% win rate)
- Advantage may diminish in bullet games due to time pressure
- Opening diversity should favor white pieces in aggressive systems

**Time Control Predictions:**
- Bullet: Higher variance, more decisive results, fewer draws
- Classical: Lower variance, more draws, deeper strategic play
- Rating correlation with performance should strengthen in longer time controls
#### Control Variables and Confounding Factors

**Player-level Controls:**
- Historical rating trajectory (improving vs. declining)
- Playing frequency and recent activity
- Specialization in specific time controls or openings
- Title status and competitive experience

**Game-level Controls:**
- Opponent strength relative to player rating
- Recent performance streak effects
- Opening preparation depth
- Hardware/software advantages (engine assistance detection)

**Environmental Controls:**
- Platform-specific features (Lichess vs. Chess.com differences)
- Time zone and circadian rhythm effects
- Internet connection stability impacts
- Mobile vs. desktop playing interface

#### Sample Size and Power Analysis

**Minimum Sample Requirements:**
- **Color analysis**: 100+ games per color for reliable estimates
- **Time control comparison**: 50+ games per category
- **Rating tier analysis**: 30+ games per 100-point rating band
- **Opening effectiveness**: 20+ games per opening family

**Power Calculations:**
- **Effect size detection**: Minimum 5% win rate difference
- **Statistical power**: Target 80% power for primary comparisons
- **Alpha level**: 0.05 with Bonferroni correction for multiple tests
- **Confidence intervals**: 95% CI for all effect estimates
