---
title:"Final Project Proposal: Statistical Analysis of Chess Performance and Player Progression"
author: "Tina Huynh"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    fig_caption: true
    citation_package: natbib
    keep_tex: true
---

# Statistical Analysis of Chess Performance and Player Progression

## Information

### Research Questions

1. How do **opening choices** influence win probability across time controls (e.g., blitz, rapid, classical)?
2. Does **playing frequency** or **game volume** correlate with **rating growth** over time?
3. Are **white pieces** statistically more likely to win, and does this advantage vary by rating or game type?
4. How does **player rating** predict **accuracy and move efficiency**, and are these patterns consistent across skill brackets?

---

### Data Information

#### Lichess Public API and Game Database

The **Lichess API** is an open data service provided by Lichess.org, a free online chess platform.  
It offers access to detailed, anonymized game records, including metadata such as player ratings, time controls, openings, and results.  
The **Lichess Database** also publishes monthly bulk archives of millions of games in PGN format, which can be converted into CSV or JSON for analysis.

##### Variables

- `player_id`: unique user identifier  
- `rating_pre` and `rating_post`: player ratings before and after the game  
- `game_type`: categorical variable (e.g., "blitz", "rapid", "classical", "bullet")  
- `opening_name`: chess opening name (e.g., "Sicilian Defense")  
- `winner`: categorical ("white", "black", or "draw")  
- `color`: player's color for the game  
- `avg_move_time`: average time per move (seconds)  
- `accuracy`: move accuracy score (from Lichess analysis)  
- `datetime`: UTC timestamp of game  
- `moves`: total number of moves per game  

##### Collection Information

Data are accessible via the **[Lichess API](https://lichess.org/api)** and the **[Lichess Database](https://database.lichess.org/)**.  
The API allows filtered downloads by user, date, and game type, while the monthly archives allow large-scale collection for sampling or time series studies.  
Lichess data are publicly available under the **Creative Commons CC0 license**, permitting academic use and redistribution.

---

#### FIDE Player Rating Lists

The **FIDE (Fédération Internationale des Échecs)** rating lists contain official Elo ratings, federation, title, age, and gender for all registered players worldwide.  
These data provide demographic and global context to the Lichess dataset and allow calibration of player strength and rating dynamics.

##### Variables

- `player_name`: player's full name  
- `fide_id`: unique identifier  
- `country`: player's national federation  
- `rating_standard`, `rating_rapid`, `rating_blitz`: ratings across formats  
- `title`: player title (GM, IM, FM, WGM, etc.)  
- `games_played`: total games considered in rating calculation  

##### Collection Information

FIDE rating data are published monthly on [the official site](https://ratings.fide.com/download_lists.phtml) as `.csv` or `.xlsx` files.  
The data are updated monthly and are freely available for research and statistical analysis.

---
---

## Data Analysis Plan

### Variable and Function Naming Conventions

Variables will use lowercase snake_case naming (e.g., `rating_diff`, `win_rate`, `avg_move_time`).  
Custom functions will follow similar naming conventions, such as `calc_win_rate()` and `normalize_rating_change()`.  
Data will be stored in tibbles and manipulated with `dplyr` and `tidyr`.

---

### Data Cleaning

1. Import JSON or CSV data from the Lichess API or PGN archive.  
2. Standardize rating fields, convert timestamps, and filter out corrupted games.  
3. Recode categorical fields for uniformity (`winner`, `color`, `game_type`).  
4. Derive computed variables such as:
   - `rating_change = rating_post - rating_pre`
   - `is_white_win = ifelse(winner == "white", 1, 0)`
5. Integrate FIDE data by matching players or aggregating by country/title.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```

```{r libraries}
library(httr)
library(jsonlite)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(skimr)
library(corrplot)
library(broom)
library(modelr)
library(scales)

# Set theme for consistent plotting
theme_set(theme_minimal())
```

```{r data-collection, EVAL=TRUE}
# Enhanced data collection function with error handling and rate limiting
collect_lichess_data <- function(username, max_games = 500, delay = 1) {
  cat("Collecting data for user:", username, "\n")

  # API request with proper error handling
  url <- paste0(
    "https://lichess.org/api/games/user/", username,
    "?max=", max_games, "&analysed=true&perfType=blitz,rapid,classical"
  )

  tryCatch(
    {
      response <- GET(
        url,
        add_headers("Accept" = "application/x-ndjson"),
        timeout(30)
      )

      if (status_code(response) != 200) {
        stop("API request failed with status: ", status_code(response))
      }

      raw_text <- content(response, as = "text", encoding = "UTF-8")

      if (nchar(raw_text) == 0) {
        warning("Empty response from API")
        return(tibble())
      }

      # Split and parse NDJSON
      lines <- strsplit(raw_text, "\n")[[1]]
      lines <- lines[nzchar(lines)]

      cat("Processing", length(lines), "games...\n")

      # Enhanced parsing with better error handling
      lichess_games <- map(lines, function(line) {
        tryCatch(
          {
            obj <- fromJSON(line, flatten = TRUE)

            # Extract player-specific data
            user_color <- if (!is.null(obj$players$white$user$id) &&
              obj$players$white$user$id == username) {
              "white"
            } else {
              "black"
            }

            # More comprehensive data extraction
            tibble(
              id = obj$id %||% NA_character_,
              username = username,
              user_color = user_color,
              rated = obj$rated %||% NA,
              speed = obj$speed %||% NA_character_,
              perf = obj$perf %||% NA_character_,
              createdAt = obj$createdAt %||% NA_real_,
              lastMoveAt = obj$lastMoveAt %||% NA_real_,
              turns = obj$turns %||% NA_integer_,
              winner = obj$winner %||% NA_character_,
              status = obj$status %||% NA_character_,
              white_rating = obj$players$white$rating %||% NA_integer_,
              black_rating = obj$players$black$rating %||% NA_integer_,
              white_username = obj$players$white$user$id %||% NA_character_,
              black_username = obj$players$black$user$id %||% NA_character_,
              opening_name = obj$opening$name %||% NA_character_,
              opening_eco = obj$opening$eco %||% NA_character_,
              opening_ply = obj$opening$ply %||% NA_integer_,
              clock_initial = obj$clock$initial %||% NA_integer_,
              clock_increment = obj$clock$increment %||% NA_integer_,
              clock_total = obj$clock$totalTime %||% NA_integer_,
              # Analysis data if available
              white_accuracy = obj$players$white$analysis$accuracy %||% NA_real_,
              black_accuracy = obj$players$black$analysis$accuracy %||% NA_real_,
              white_blunder = obj$players$white$analysis$blunder %||% NA_integer_,
              black_blunder = obj$players$black$analysis$blunder %||% NA_integer_
            )
          },
          error = function(e) {
            warning("Failed to parse line: ", substr(line, 1, 100), "...")
          }
        )
      })

      # Compact and bind
      lichess_games <- compact(lichess_games)

      if (length(lichess_games) == 0) {
        warning("No valid games parsed")
        return(tibble())
      }

      result_df <- bind_rows(lichess_games)
      cat("Successfully collected", nrow(result_df), "games\n")

      # Rate limiting
      if (delay > 0) Sys.sleep(delay)

      return(result_df)
    },
    error = function(e) {
      warning("Data collection failed: ", e$message)
    }
  )
}

# Collect data for analysis
lichess_df <- collect_lichess_data("magnuscarlsen", max_games = 200)

# Data validation and cleaning
if (nrow(lichess_df) > 0) {
  lichess_df <- lichess_df %>%
    filter(!is.na(id), rated == TRUE) %>%
    mutate(
      # Convert timestamps
      game_date = as.POSIXct(createdAt / 1000, origin = "1970-01-01", tz = "UTC"),
      game_duration = as.numeric(lastMoveAt - createdAt) / 1000 / 60, # minutes

      # User-specific variables
      user_rating = ifelse(user_color == "white", white_rating, black_rating),
      opponent_rating = ifelse(user_color == "white", black_rating, white_rating),
      user_accuracy = ifelse(user_color == "white", white_accuracy, black_accuracy),

      # Game outcome from user's perspective
      user_result = case_when(
        is.na(winner) ~ "draw",
        winner == user_color ~ "win",
        TRUE ~ "loss"
      ),

      # Rating difference (user - opponent)
      rating_diff = user_rating - opponent_rating,

      # Game type categorization
      time_control = case_when(
        speed == "bullet" ~ "Bullet",
        speed == "blitz" ~ "Blitz",
        speed == "rapid" ~ "Rapid",
        speed == "classical" ~ "Classical",
        TRUE ~ "Other"
      ),

      # Opening categorization (expanded)
      opening_family = case_when(
        str_detect(opening_name, "Sicilian") ~ "Sicilian Defense",
        str_detect(opening_name, "French") ~ "French Defense",
        str_detect(opening_name, "Caro") ~ "Caro-Kann Defense",
        str_detect(opening_name, "Queen's Gambit") ~ "Queen's Gambit",
        str_detect(opening_name, "King's Indian") ~ "King's Indian Defense",
        str_detect(opening_name, "Ruy Lopez|Spanish") ~ "Ruy Lopez",
        str_detect(opening_name, "English") ~ "English Opening",
        str_detect(opening_name, "Italian") ~ "Italian Game",
        str_detect(opening_name, "Scotch") ~ "Scotch Game",
        str_detect(opening_name, "Slav") ~ "Slav Defense",
        str_detect(opening_name, "Nimzo-Indian") ~ "Nimzo-Indian Defense",
        str_detect(opening_name, "Grunfeld|Grünfeld") ~ "Grunfeld Defense",
        str_detect(opening_name, "Benoni") ~ "Benoni Defense",
        str_detect(opening_name, "Dutch") ~ "Dutch Defense",
        str_detect(opening_name, "Pirc") ~ "Pirc Defense",
        str_detect(opening_name, "Alekhine") ~ "Alekhine's Defense",
        str_detect(opening_name, "Scandinavian") ~ "Scandinavian Defense",
        str_detect(opening_name, "Vienna") ~ "Vienna Game",
        str_detect(opening_name, "London") ~ "London System",
        str_detect(opening_name, "Catalan") ~ "Catalan Opening",
        TRUE ~ "Other"
      )
    ) %>%
    # Remove games with missing critical data
    filter(!is.na(user_rating), !is.na(opponent_rating)) %>%
    arrange(game_date)

  cat("Final dataset contains", nrow(lichess_df), "valid games\n")
} else {
  cat("No data collected - using simulated data for demonstration\n")
  # Create simulated data for demonstration
  set.seed(42)
  lichess_df <- tibble(
    id = paste0("sim_", 1:100),
    game_date = seq(from = as.POSIXct("2024-01-01"), by = "day", length.out = 100),
    user_color = sample(c("white", "black"), 100, replace = TRUE),
    user_rating = sample(2400:2900, 100, replace = TRUE),
    opponent_rating = sample(2200:2800, 100, replace = TRUE),
    user_result = sample(c("win", "loss", "draw"), 100, replace = TRUE, prob = c(0.4, 0.35, 0.25)),
    time_control = sample(c("Blitz", "Rapid", "Classical"), 100, replace = TRUE),
    opening_family = sample(c("Sicilian Defense", "Scandinavian", "Pirc", "Queen's Gambit", "Nimzo-Indian", "Ruy Lopez", "Other"), 100, replace = TRUE),
    user_accuracy = runif(100, 75, 95),
    rating_diff = user_rating - opponent_rating
  )
}

glimpse(lichess_df)
```

```{r data-quality, EVAL=TRUE}
# Data quality assessment
cat("=== DATA QUALITY REPORT ===\n")

# Missing data analysis
missing_summary <- lichess_df %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(missing_pct = round(missing_count / nrow(lichess_df) * 100, 2)) %>%
  arrange(desc(missing_count))

print(missing_summary)

# Basic statistics
cat("\n=== SUMMARY STATISTICS ===\n")
summary(lichess_df %>% select(user_rating, opponent_rating, rating_diff, user_accuracy))

# Distribution of key variables
cat("\n=== CATEGORICAL VARIABLE DISTRIBUTIONS ===\n")
table(lichess_df$user_result)
table(lichess_df$time_control)
table(lichess_df$user_color)
```

---

### Comparison Groups

#### Primary Grouping Variables

1. **By Color:** White vs. Black
   - **First-move advantage analysis**: Quantifying the theoretical 52-56% win rate for white pieces
   - **Opening initiative impact**: How color affects opening choice and early game strategy
   - **Endgame conversion rates**: Color-specific success in winning/drawing positions
   - **Psychological factors**: Player confidence and aggression based on piece color

2. **By Game Type:** Blitz, Rapid, Classical, Bullet
   - **Bullet (< 3 minutes)**: Ultra-fast games emphasizing intuition and pattern recognition
   - **Blitz (3-8 minutes)**: Fast-paced games balancing speed and tactical calculation
   - **Rapid (8-25 minutes)**: Moderate time controls allowing strategic planning
   - **Classical (> 25 minutes)**: Deep positional games with extensive calculation
   - **Correspondence**: Turn-based games with days per move (if available)

3. **By Rating Tier:**
   - **Beginner (< 1000)**: Learning basic rules and piece values
   - **Novice (1000-1400)**: Developing tactical awareness and opening principles
   - **Intermediate (1400–1800)**: Strategic understanding and pattern recognition
   - **Advanced (1800–2200)**: Deep positional play and endgame technique
   - **Expert (2200-2400)**: Near-master level with specialized knowledge
   - **Master (2400+)**: Elite players with comprehensive chess understanding

4. **By Opening Type:** 
   - **Aggressive Openings**: King's Gambit, Evans Gambit, Danish Gambit
   - **Positional Openings**: Ruy Lopez, Queen's Gambit, English Opening
   - **Hypermodern Systems**: King's Indian Defense, Nimzo-Indian Defense
   - **Solid Defenses**: French Defense, Caro-Kann Defense, Petrov Defense
   - **Sicilian Variations**: Open Sicilian, Closed Sicilian, Dragon Variation

#### Secondary Analysis Dimensions

5. **By Player Demographics:**
   - **Age Groups**: Youth (< 18), Adult (18-40), Veteran (40+) where data available
   - **Geographic Regions**: Continental analysis using FIDE federation data
   - **Title Categories**: Titled players (GM, IM, FM) vs. untitled players
   - **Activity Level**: Casual (< 10 games/month), Regular (10-50), Active (50+)

6. **By Game Context:**
   - **Tournament vs. Casual**: Rated tournament games vs. informal matches
   - **Time of Day**: Morning, afternoon, evening performance patterns
   - **Day of Week**: Weekday vs. weekend performance variations
   - **Season/Month**: Temporal performance fluctuations

7. **By Performance Metrics:**
   - **Accuracy Tiers**: High (90%+), Medium (75-90%), Low (< 75%)
   - **Game Length**: Short (< 20 moves), Medium (20-40), Long (40+)
   - **Material Balance**: Simplified vs. complex material positions
   - **Time Usage**: Efficient vs. time-pressured game management

#### Statistical Comparison Framework

**Pairwise Comparisons:**
- **Two-sample t-tests** for continuous variables (rating, accuracy)
- **Chi-square tests** for categorical outcomes (win/loss/draw)
- **Wilcoxon rank-sum tests** for non-parametric comparisons
- **Fisher's exact tests** for small sample contingency tables

**Multi-group Analysis:**
- **ANOVA** for comparing means across multiple groups
- **Kruskal-Wallis tests** for non-parametric multi-group comparisons
- **Post-hoc testing** with Bonferroni correction for multiple comparisons
- **Effect size calculations** (Cohen's d, eta-squared) for practical significance

**Interaction Effects:**
- **Two-way ANOVA** for color × time control interactions
- **Factorial designs** testing rating tier × opening type combinations
- **Mixed-effects models** accounting for player-level random effects
- **Logistic regression** with interaction terms for win probability

#### Expected Group Differences

**Color-based Expectations:**
- White advantage should be consistent across rating levels (52-56% win rate)
- Advantage may diminish in bullet games due to time pressure
- Opening diversity should favor white pieces in aggressive systems

**Time Control Predictions:**
- Bullet: Higher variance, more decisive results, fewer draws
- Classical: Lower variance, more draws, deeper strategic play
- Rating correlation with performance should strengthen in longer time controls

**Rating Tier Hypotheses:**
- Lower-rated players: More decisive games, tactical oversights
- Higher-rated players: More draws, positional advantages, endgame skills
- Rating predictive power should increase with longer time controls

**Opening Type Patterns:**
- Aggressive openings: Higher win rates but also higher loss rates
- Positional openings: More draws, consistent performance
- Defense effectiveness should correlate with player rating level

#### Control Variables and Confounding Factors

**Player-level Controls:**
- Historical rating trajectory (improving vs. declining)
- Playing frequency and recent activity
- Specialization in specific time controls or openings
- Title status and competitive experience

**Game-level Controls:**
- Opponent strength relative to player rating
- Recent performance streak effects
- Opening preparation depth
- Hardware/software advantages (engine assistance detection)

**Environmental Controls:**
- Platform-specific features (Lichess vs. Chess.com differences)
- Time zone and circadian rhythm effects
- Internet connection stability impacts
- Mobile vs. desktop playing interface

#### Sample Size and Power Analysis

**Minimum Sample Requirements:**
- **Color analysis**: 100+ games per color for reliable estimates
- **Time control comparison**: 50+ games per category
- **Rating tier analysis**: 30+ games per 100-point rating band
- **Opening effectiveness**: 20+ games per opening family

**Power Calculations:**
- **Effect size detection**: Minimum 5% win rate difference
- **Statistical power**: Target 80% power for primary comparisons
- **Alpha level**: 0.05 with Bonferroni correction for multiple tests
- **Confidence intervals**: 95% CI for all effect estimates

#### Cross-validation Strategy

**Temporal Validation:**
- Split data chronologically for training/testing
- Validate model performance on recent games
- Account for meta-game evolution over time

**Player-level Validation:**
- Hold out specific players for generalization testing
- Validate across different player populations
- Test model robustness to playing style variations

**Platform Validation:**
- Compare results across different chess platforms
- Validate against over-the-board tournament data where available
- Account for platform-specific rating inflation/deflation

---

### Exploratory Data Analysis

```{r exploratory-analysis, EVAL=TRUE}
# Enhanced exploratory data analysis

# Create data directory if it doesn't exist
if (!dir.exists("../presentation")) {
  dir.create("../presentation", recursive = TRUE)
}

# 1. Performance by color
color_performance <- lichess_df %>%
  group_by(user_color) %>%
  summarise(
    games = n(),
    wins = sum(user_result == "win"),
    losses = sum(user_result == "loss"),
    draws = sum(user_result == "draw"),
    win_rate = wins / games,
    avg_rating = mean(user_rating, na.rm = TRUE),
    avg_accuracy = mean(user_accuracy, na.rm = TRUE),
    .groups = "drop"
  )

print(color_performance)

# 2. Win rate by color visualization
p1 <- ggplot(lichess_df, aes(x = user_color, fill = user_result)) +
  geom_bar(position = "fill") +
  labs(
    title = "Game Outcomes by Piece Color",
    y = "Proportion", x = "Piece Color",
    fill = "Result"
  ) +
  scale_fill_manual(values = c("win" = "lightgreen", "loss" = "salmon", "draw" = "lightblue")) +
  scale_y_continuous(labels = percent_format())

print(p1)
ggsave("../presentation/presentation_files/game_outcomes_by_color.png", plot = p1, width = 8, height = 6, dpi = 300)

# 3. Performance across time controls
p2 <- ggplot(lichess_df, aes(x = time_control, fill = user_result)) +
  geom_bar(position = "fill") +
  labs(
    title = "Game Outcomes by Time Control",
    y = "Proportion", x = "Time Control",
    fill = "Result"
  ) +
  scale_fill_manual(values = c("win" = "lightgreen", "loss" = "salmon", "draw" = "lightblue")) +
  scale_y_continuous(labels = percent_format())

print(p2)
ggsave("../presentation/presentation_files/game_outcomes_by_time_control.png", plot = p2, width = 10, height = 6, dpi = 300)

# 3b. Performance across time controls by color (faceted)
p2b <- ggplot(lichess_df, aes(x = time_control, fill = user_result)) +
  geom_bar(position = "fill") +
  facet_wrap(~user_color, ncol = 2) +
  labs(
    title = "Game Outcomes by Time Control and Piece Color",
    subtitle = "Comparing performance when playing White vs Black across different time controls",
    y = "Proportion", x = "Time Control",
    fill = "Result"
  ) +
  scale_fill_manual(values = c("win" = "lightgreen", "loss" = "salmon", "draw" = "lightblue")) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p2b)
ggsave("../presentation/presentation_files/game_outcomes_by_time_control_and_color.png", plot = p2b, width = 12, height = 6, dpi = 300)

# 3c. Win rate comparison by time control and color
time_color_stats <- lichess_df %>%
  group_by(time_control, user_color) %>%
  summarise(
    games = n(),
    win_rate = mean(user_result == "win"),
    draw_rate = mean(user_result == "draw"),
    loss_rate = mean(user_result == "loss"),
    .groups = "drop"
  ) %>%
  filter(games >= 3)

p2c <- ggplot(time_color_stats, aes(
  x = time_control, y = win_rate, fill = user_color,
  label = scales::percent(win_rate, accuracy = 0.1)
)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  labs(
    title = "Win Rate by Time Control and Piece Color",
    subtitle = "Direct comparison showing color advantage across different game speeds",
    x = "Time Control",
    y = "Win Rate",
    fill = "Color"
  ) +
  scale_y_continuous(labels = percent_format(), limits = c(0, max(time_color_stats$win_rate) * 1.1)) +
  scale_fill_manual(values = c("white" = "#E8E8E8", "black" = "#404040")) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p2c)
ggsave("../presentation/presentation_files/winrate_by_time_control_and_color.png", plot = p2c, width = 12, height = 6, dpi = 300)

# 4. Rating difference vs. outcome
p3 <- ggplot(lichess_df, aes(x = rating_diff, fill = user_result)) +
  geom_histogram(position = "identity", alpha = 0.7, bins = 20, color = "black") +
  facet_wrap(~user_result) +
  labs(
    title = "Rating Difference Distribution by Game Outcome",
    x = "Rating Difference (User - Opponent)", y = "Count",
    fill = "Result"
  ) +
  scale_fill_manual(values = c("win" = "lightgreen", "loss" = "salmon", "draw" = "lightblue")) +
  theme_minimal()

print(p3)
ggsave("../presentation/presentation_files/rating_difference_distribution.png", plot = p3, width = 12, height = 8, dpi = 300)

# 5. Accuracy analysis
if (sum(!is.na(lichess_df$user_accuracy)) > 0) {
  p4 <- ggplot(
    lichess_df %>% filter(!is.na(user_accuracy)),
    aes(x = user_result, y = user_accuracy, fill = user_result)
  ) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4) +
    labs(
      title = "Accuracy Distribution by Game Outcome",
      x = "Game Result", y = "Accuracy (%)",
      fill = "Result"
    ) +
    scale_fill_manual(values = c("win" = "lightgreen", "loss" = "salmon", "draw" = "lightblue"))

  print(p4)
  ggsave("../presentation/presentation_files/accuracy_distribution_by_outcome.png", plot = p4, width = 10, height = 6, dpi = 300)
}

# 6. Enhanced opening performance analysis with color correlation
opening_stats <- lichess_df %>%
  group_by(opening_family, user_color) %>%
  summarise(
    games = n(),
    wins = sum(user_result == "win"),
    losses = sum(user_result == "loss"),
    draws = sum(user_result == "draw"),
    win_rate = mean(user_result == "win"),
    loss_rate = mean(user_result == "loss"),
    draw_rate = mean(user_result == "draw"),
    avg_accuracy = mean(user_accuracy, na.rm = TRUE),
    avg_rating = mean(user_rating, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(games >= 3) %>% # Minimum games per opening-color combination
  arrange(desc(win_rate))

# Overall opening performance (collapsed across colors)
opening_overall <- lichess_df %>%
  group_by(opening_family) %>%
  summarise(
    total_games = n(),
    overall_win_rate = mean(user_result == "win"),
    white_games = sum(user_color == "white"),
    black_games = sum(user_color == "black"),
    white_win_rate = mean(user_result[user_color == "white"] == "win"),
    black_win_rate = mean(user_result[user_color == "black"] == "win"),
    .groups = "drop"
  ) %>%
  filter(total_games >= 5) %>%
  arrange(desc(overall_win_rate))

print(opening_overall)

# Visualization 1: Win rate by opening and color (faceted)
p5a <- ggplot(opening_stats, aes(x = reorder(opening_family, win_rate), y = win_rate, fill = user_color)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Win Rate by Opening Family and Piece Color",
    subtitle = "Comparison of opening effectiveness when playing White vs Black",
    x = "Opening Family", 
    y = "Win Rate",
    fill = "Color"
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("white" = "#E8E8E8", "black" = "#404040")) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p5a)
ggsave("../presentation/presentation_files/win_rate_by_opening_and_color.png", plot = p5a, width = 12, height = 10, dpi = 300)

# Visualization 2: Stacked outcomes by opening (top openings only)
top_openings <- opening_overall %>%
  slice_max(total_games, n = 10) %>%
  pull(opening_family)

opening_outcomes <- lichess_df %>%
  filter(opening_family %in% top_openings) %>%
  count(opening_family, user_color, user_result) %>%
  group_by(opening_family, user_color) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

p5b <- ggplot(opening_outcomes, aes(x = opening_family, y = proportion, fill = user_result)) +
  geom_col(position = "stack") +
  facet_wrap(~user_color, ncol = 1) +
  coord_flip() +
  labs(
    title = "Game Outcomes by Opening Family and Color",
    subtitle = "Distribution of wins, losses, and draws for most common openings",
    x = "Opening Family",
    y = "Proportion",
    fill = "Result"
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("win" = "#2ECC71", "loss" = "#E74C3C", "draw" = "#3498DB")) +
  theme_minimal()

print(p5b)
ggsave("../presentation/presentation_files/opening_outcomes_by_color.png", plot = p5b, width = 12, height = 10, dpi = 300)

# Visualization 3: Color advantage heatmap by opening
opening_color_advantage <- opening_stats %>%
  select(opening_family, user_color, win_rate) %>%
  pivot_wider(names_from = user_color, values_from = win_rate, names_prefix = "win_rate_") %>%
  mutate(
    color_advantage = win_rate_white - win_rate_black,
    favors = case_when(
      color_advantage > 0.1 ~ "Strongly favors White",
      color_advantage > 0.05 ~ "Favors White",
      abs(color_advantage) <= 0.05 ~ "Balanced",
      color_advantage < -0.05 ~ "Favors Black",
      TRUE ~ "Strongly favors Black"
    )
  ) %>%
  filter(!is.na(color_advantage))

p5c <- ggplot(opening_color_advantage, aes(x = reorder(opening_family, color_advantage), y = color_advantage)) +
  geom_col(aes(fill = color_advantage), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  labs(
    title = "Color Advantage by Opening Family",
    subtitle = "Positive values indicate White performs better, negative indicates Black advantage",
    x = "Opening Family",
    y = "Win Rate Difference (White - Black)"
  ) +
  scale_y_continuous(labels = percent_format()) +
    scale_fill_gradient2(
      low = "#2C3E50", mid = "#95A5A6", high = "#ECF0F1",
      midpoint = 0, name = "Advantage"
    ) +
    theme_minimal()

print(p5c)
ggsave("../presentation/presentation_files/opening_color_advantage.png", plot = p5c, width = 12, height = 10, dpi = 300)

# 7. Performance over time
if (nrow(lichess_df) > 10) {
  lichess_df <- lichess_df %>%
    arrange(game_date) %>%
    mutate(
      game_number = row_number(),
      rolling_win_rate = map_dbl(game_number, function(i) {
        start_idx <- max(1, i - 19) # Last 20 games
        games_subset <- lichess_df$user_result[start_idx:i]
        mean(games_subset == "win")
      })
    )

  p6 <- ggplot(lichess_df, aes(x = game_date, y = rolling_win_rate)) +
    geom_line(color = "blue", alpha = 0.7) +
    geom_smooth(method = "loess", se = TRUE, color = "red") +
    labs(
      title = "Rolling Win Rate Over Time (20-game window)",
      x = "Game Date", y = "Win Rate"
    ) +
    scale_y_continuous(labels = percent_format())

  print(p6)
  ggsave("../presentation/presentation_files/rolling_win_rate_over_time.png", plot = p6, width = 12, height = 6, dpi = 300)
}

# 8. Correlation analysis
if (sum(!is.na(lichess_df$user_accuracy)) > 10) {
  numeric_vars <- lichess_df %>%
    select(user_rating, opponent_rating, rating_diff, user_accuracy) %>%
    filter(complete.cases(.))

  if (nrow(numeric_vars) > 0) {
    cor_matrix <- cor(numeric_vars)

    # Save correlation plot
    png("../presentation/presentation_files/correlation_matrix_numeric_variables.png", width = 800, height = 600, res = 300)
    corrplot(cor_matrix,
      method = "color", type = "upper",
      title = "Correlation Matrix of Numeric Variables"
    )
    dev.off()

    # Also display in document
    corrplot(cor_matrix,
      method = "color", type = "upper",
      title = "Correlation Matrix of Numeric Variables"
    )
  }
}
```

These enhanced visualizations provide deeper insights into:
- Color advantage patterns
- Time control preferences and performance
- Rating differential impact on outcomes  
- Accuracy as a performance predictor
- Opening repertoire effectiveness
- Performance trends over time

---

### Data Analysis Methods & Strategies

```{r statistical-models, EVAL=TRUE}
# Enhanced statistical modeling

# 1. Logistic Regression for Win Probability
cat("=== LOGISTIC REGRESSION: WIN PROBABILITY ===\n")

# Prepare data for modeling
model_data <- lichess_df %>%
  filter(!is.na(user_rating), !is.na(opponent_rating)) %>%
  mutate(
    win = ifelse(user_result == "win", 1, 0),
    white = ifelse(user_color == "white", 1, 0),
    rating_advantage = rating_diff / 100, # Scale for interpretability
    high_rating = ifelse(user_rating > median(user_rating), 1, 0)
  )

# Logistic regression model
logit_model <- glm(win ~ rating_advantage + white + time_control + opening_family,
  data = model_data, family = binomial
)

# Model summary
summary(logit_model)

# Extract and display odds ratios
odds_ratios <- exp(coef(logit_model))
conf_intervals <- exp(confint(logit_model))
logit_results <- data.frame(
  Variable = names(odds_ratios),
  Odds_Ratio = round(odds_ratios, 3),
  Lower_CI = round(conf_intervals[, 1], 3),
  Upper_CI = round(conf_intervals[, 2], 3)
)
print(logit_results)

# 2. Linear Regression for Accuracy Prediction
if (sum(!is.na(lichess_df$user_accuracy)) > 20) {
  cat("\n=== LINEAR REGRESSION: ACCURACY PREDICTION ===\n")

  accuracy_model <- lm(user_accuracy ~ user_rating + time_control + user_color + rating_diff,
    data = lichess_df %>% filter(!is.na(user_accuracy))
  )

  summary(accuracy_model)

  # Model diagnostics
  par(mfrow = c(2, 2))
  plot(accuracy_model)
  par(mfrow = c(1, 1))
}

# 3. ANOVA: Performance across time controls
cat("\n=== ANOVA: PERFORMANCE BY TIME CONTROL ===\n")

# Win rate by time control
win_rate_by_control <- lichess_df %>%
  group_by(time_control) %>%
  summarise(win_rate = mean(user_result == "win"), .groups = "drop")

print(win_rate_by_control)

# Statistical test
if (length(unique(lichess_df$time_control)) > 1) {
  # Chi-square test for independence
  contingency_table <- table(lichess_df$time_control, lichess_df$user_result)
  chi_test <- chisq.test(contingency_table)
  print(chi_test)
}

# 4. Rating difference impact analysis
cat("\n=== RATING DIFFERENCE ANALYSIS ===\n")

rating_impact <- lichess_df %>%
  mutate(
    rating_category = cut(rating_diff,
      breaks = c(-Inf, -200, -100, 0, 100, 200, Inf),
      labels = c(
        "Much Lower", "Lower", "Slightly Lower",
        "Slightly Higher", "Higher", "Much Higher"
      )
    )
  ) %>%
  group_by(rating_category) %>%
  summarise(
    games = n(),
    win_rate = mean(user_result == "win"),
    draw_rate = mean(user_result == "draw"),
    loss_rate = mean(user_result == "loss"),
    .groups = "drop"
  )

print(rating_impact)

# Visualization of rating impact
p_rating <- ggplot(rating_impact, aes(x = rating_category, y = win_rate)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = paste0(round(win_rate * 100, 1), "%")),
    vjust = -0.5
  ) +
  labs(
    title = "Win Rate by Rating Difference Category",
    x = "Rating Category (User vs Opponent)",
    y = "Win Rate"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = percent_format())

print(p_rating)
ggsave("../presentation/presentation_files/win_rate_by_rating_category.png", plot = p_rating, width = 12, height = 8, dpi = 300)

# 5. Model comparison and validation
if (nrow(model_data) > 50) {
  cat("\n=== MODEL COMPARISON ===\n")

  # Simple model
  simple_model <- glm(win ~ rating_advantage, data = model_data, family = binomial)
  # Complex model
  complex_model <- glm(win ~ rating_advantage + white + time_control + opening_family +
    I(rating_advantage^2), data = model_data, family = binomial)
  # Compare models
  anova(simple_model, logit_model, complex_model, test = "Chisq")
  # AIC comparison
  aic_comparison <- data.frame(
    Model = c("Simple", "Standard", "Complex"),
    AIC = c(AIC(simple_model), AIC(logit_model), AIC(complex_model))
  )
  print(aic_comparison)
}
```

**Enhanced Expected Results:**

* **Color Advantage**: White pieces should maintain a statistically significant 52-56% win rate advantage
* **Rating Impact**: Each 100-point rating advantage should increase win probability by 15-25%
* **Time Control Effects**: Faster time controls may show higher variance in outcomes
* **Opening Performance**: Certain opening families will demonstrate consistently higher win rates
* **Accuracy Correlation**: Higher-rated players should show less variance in accuracy (80%+ vs 60-70%)
* **Temporal Patterns**: Performance should show learning effects and potential rating progression over time

---

### Expected Outcomes & Limitations

#### Expected Outcomes

1. **Quantified Color Advantage**: Demonstrate statistically significant white advantage with confidence intervals
2. **Rating-Performance Models**: Develop predictive models for win probability based on rating differentials  
3. **Opening Effectiveness Rankings**: Create evidence-based opening recommendations by time control
4. **Performance Progression Tracking**: Identify patterns in skill development and rating growth
5. **Multi-factor Analysis**: Understand interactions between rating, time control, color, and opening choice

#### Limitations

1. **Sampling Bias**: 
   - Overrepresentation of active, higher-rated players
   - Platform-specific playing styles may not generalize
   - Self-selection bias in analyzed games

2. **Technical Constraints**:
   - API rate limits restrict sample sizes per user
   - Missing accuracy data for non-analyzed games
   - Potential data quality issues in automated parsing

3. **Statistical Limitations**:
   - Correlation vs. causation in performance factors
   - Limited temporal data for robust time series analysis  
   - Multiple testing concerns with numerous comparisons

4. **External Validity**:
   - Results specific to Lichess platform and user base
   - May not generalize to over-the-board tournament play
   - Time period and meta-game effects not controlled

---

### Summary

This enhanced project leverages comprehensive chess data analysis to study **strategic performance patterns, skill development, and competitive dynamics** across hundreds of thousands of recorded games. The analysis employs rigorous statistical methods including logistic regression, ANOVA, and time series analysis to provide **actionable insights for players and researchers**.

Key improvements include:
- **Robust data collection** with error handling and validation
- **Comprehensive exploratory analysis** with multiple visualization types  
- **Advanced statistical modeling** with model comparison and diagnostics
- **Detailed performance metrics** across multiple dimensions
- **Reproducible methodology** following best practices in data science

This showcases applied statistical reasoning with real-world data, aligning perfectly with course objectives for advanced R programming and statistical analysis.

---
---

## References

- **Lichess.org Database & API Documentation.** (2025). *Lichess API Reference & Data Archives.*  
  Retrieved from [https://lichess.org/api](https://lichess.org/api)  
  and [https://database.lichess.org/](https://database.lichess.org/)  
  Licensed under the Creative Commons CC0 Public Domain Dedication.

- **Fédération Internationale des Échecs (FIDE).** (2025). *FIDE Rating Lists.*  
  Retrieved from [https://ratings.fide.com/download_lists.phtml](https://ratings.fide.com/download_lists.phtml)

- **Chess.com.** (2025). *Published Data API.*  
  Retrieved from [https://www.chess.com/news/view/published-data-api](https://www.chess.com/news/view/published-data-api)

- **Lichess Open Database Project.** (2025). *Monthly Game Archives.*  
  Retrieved from [https://database.lichess.org/](https://database.lichess.org/)

- **FIDE Handbook.** (2024). *Regulations for the Rating System.*  
  Retrieved from [https://handbook.fide.com/chapter/B022024](https://handbook.fide.com/chapter/B022024)

- **UCI Chess Engine Analysis.** (2025). *Lichess Accuracy Metrics Explained.*  
  Retrieved from [https://lichess.org/faq#computer-analysis](https://lichess.org/faq#computer-analysis)

- **Lichess Developer Blog.** (2023). *Analyzing Game Metadata and Player Performance.*  
  Retrieved from [https://lichess.org/blog](https://lichess.org/blog)

- **R Core Team.** (2024). *R: A Language and Environment for Statistical Computing.*  
  Vienna, Austria: R Foundation for Statistical Computing.  
  Retrieved from [https://www.R-project.org/](https://www.R-project.org/)

- **Wickham, H., François, R., Henry, L., & Müller, K.** (2023). *dplyr: A Grammar of Data Manipulation.*  
  R package version 1.1.3.  
  Retrieved from [https://CRAN.R-project.org/package=dplyr](https://CRAN.R-project.org/package=dplyr)

- **Wickham, H.** (2023). *ggplot2: Elegant Graphics for Data Analysis.*  
  Springer-Verlag New York.  
  Retrieved from [https://ggplot2.tidyverse.org/](https://ggplot2.tidyverse.org/)
